% SPDX-License-Identifier: CC-BY-4.0
%
% Copyright (c) 2023 Nelson Vieira
%
% @author Nelson Vieira <nelson0.vieira@gmail.com>
% @license CC-BY-4.0 <https://creativecommons.org/licenses/by/4.0/legalcode.txt>
\chapter*{Appendix}

Appendix content...

\section*{Survey}
\label{appendix:survey}

\subsection*{Section 1: General knowledge and attitudes towards privacy}

The purpose of this section is to ask general questions about your knowledge
in the area of information privacy.

1. "Privacy is important to me". Do you agree with this statement?

2. "I know techniques to guarantee privacy and the protection of my data when I use the Internet". Do you agree with this statement?

3. Could you give examples of techniques or strategies you use to guarantee the privacy and protection of your data?


4. Do you think your data privacy is important?


5. How do you define data (or digital) privacy?


6. Do you think your data privacy is relevant nowadays?


7. "Data privacy is a human right". Do you agree with this statement?


8. "Data privacy is a consumer right". Do you agree with this statement?


9. In your opinion, are security and privacy synonymous?


10. How familiar are you with the following terms?


11. Do you think your data is stored/collected when you use the internet?


12. What means do you think are used to collect your data?


13. What kind of information can be extracted using the collected data?


14. For what purposes can organizations use your data?


15. Which organizations do you consider to have the best privacy behaviour?


16. Which organizations do you consider to have the worst privacy behaviour?


17. During your day-to-day life, how often do you use your phone to access the internet?


18. "I am concerned about my privacy when using my mobile phone when accessing the internet". Do you agree with this statement?


19. "I consider that accessing the internet through my phone is safer than through a computer". Do you agree with this statement?


20. "I try to block the collection of data from applications installed on my phone". Do you agree with this statement?


21. When using a website, do you usually read the Privacy Policy?


22. How often do you allow the use of cookies?


23. When you accept / become aware of the cookies policy, why do you take this decision?


24. Are you aware of the concept of "profiling" or "automated processing of personal information"?


25. Do you consider that your internet activity contributes to the development of profiling?


26. Are you aware of regulations such as the General Data Protection Regulation or the California Consumer Privacy Act?


27. If you answered yes, how did you learn about the 'General Data Protection Regulation' or the 'California Consumer Privacy Act'?


28. Are you interested in finding out more about regulations or legislation related to digital privacy?


29. According to the General Data Protection Regulation, personal data is:


30. "My data was more protected after the implementation of regulations such as the General Data Protection Regulation". Do you agree with this statement?

\subsection*{Section 2: Disposition for sharing personal information}

This section is intended to ask questions to generally understand your willingness to share personal information.

1. What kind of personal information are you willing to share at any time?

2. In what situations are you willing to provide more personal information?

3. Do you agree to share health data that can identify you with health professionals?


4. Do you agree to share health data that cannot identify you with health professionals?

5. What kind of applications do you have installed on your smartphone?

6. Before sharing your data, do you consult any of the following information?

7. How often do you find privacy policies?

8. Are you aware of the duties of a Data Protection Officer (DPO)?

9. "I am interested in knowing where and how my personal information is used." Do you agree with this statement?

10. "I am not familiar with the purpose of data collection but would like to know more". Do you agree with this statement?

11. "The length (or number of words) of the privacy notice affects my willingness to read it." Do you agree with this statement?

12. "The font size of the privacy notice affects my willingness to read it." Do you agree with this statement?

13. "Usually, I'm afraid I won't be able to use a product or service if I don't agree with the privacy notice." Do you agree with this statement?

14. "I don't need to read the privacy notice if I trust the institution". Do you agree with this statement?

\subsection*{Section 3: Privacy concerns}

This section aims to ask questions related to possible fears associated with sharing personal information.


1. How concerned are you about organizations collecting and using your online activity?

2. How concerned are you about organizations sharing your data with third parties?

3. How concerned are you about organizations tracking your online behavior and thus obtaining your personal data?

4. How concerned are you about public institutions or intelligence services analyzing your online movements?

5. How concerned are you that other people obtain your personal data without your consent?

6. How concerned are you that other people find information about you online?


7. How concerned are you that other people are disclosing information about you without your knowledge?

8. How concerned are you about other people sharing your personal data (photos, address, mobile phone number, etc.) with others without your consent?

9. How concerned are you that other people publish your personal data (photos, address, mobile phone number, etc.) on the internet without your consent?

10. How concerned are you about an unknown person claiming to be you on the internet?

11. How concerned are you about the possibility that someone may misuse your identity on the internet?

\subsection*{Section 4: Current online habits and practices}

This section brings together general questions about using the Internet in your daily life.


1. Do you have internet access at home?

2. How often do you use the internet?

3. How much time do you spend per day surfing the internet?

4. What device(s) do you use to access the internet?

5. For what purposes do you use the internet?

6. What platforms/applications do you use in your day-to-day life?

7. On social media, do you usually:

8. Your activity on social media goes through:

\subsection*{Section 5: Profile identification}

This section brings together more specific questions about using profiles to make it easier to create personalized experiences.


1. Do you know the concept of profiling?

2. If a law enforcement officer asks for your personal data, are you willing to share it?

3. In what situations are you willing to provide personal data when data collection is disclosed?

4. In what situations are you willing to provide personal data when data collection is not disclosed?


5. Do you usually answer online questionnaires when they are requested?

6. When answering questionnaires, do you enter any false/incorrect information?

7. When creating an account on an online platform, have you ever entered false personal data?

7.1. If yes, why did you enter this false data?

7.2. When you have to enter false data, what types of false data do you enter?

8. In your opinion, does the data you disclose on one platform influence the use of another? Justify.

9. Do you think companies sell your personal data? Justify.

10. Can the data I disclose serve to create a profile of my online habits?

11. "The information I disclose on the internet can serve to identify me". Do you agree with this statement?

12. Are you aware of data brokers?

13. If so, explain what data brokers are and what they do.

\subsection*{Section 6: Knowledge and habits regarding the Internet of Things}


1. What do you understand by Internet of Things?

2. Have you ever used an Internet of Things device (smartwatch, contactless cards, air or sea traffic applications)?

3. Do you have an Internet of Things device in your home (assistants like Alexa, smart lock, video surveillance, etc.)?

3.1. If so, how often do you interact with that device(s)?

3.2. And for what purposes do you use the device(s)?

3.3. Why did you feel the need to have an Internet of Things device?

4. How familiar are you with the following terms?

5. In your opinion, how many Internet of Things devices are there today?

\subsection*{Demographic data}

We thank you in advance for your participation. And we ask that you fill in
the following questions with demographic information. This section intends
to collect general information that allows to characterize the users in
statistical terms that will be part of this study when sharing their personal
perceptions in terms of privacy.

- Age

- Genre

- Country

- District / State

- Level of Education

- Professional area

- Annual income



\section*{Software Requirements Specification}
\label{appendix:a}

Content...

\section*{Usability Tests}
\label{appendix:usability_tests}

Content...
