% SPDX-License-Identifier: CC-BY-4.0
%
% Copyright (c) 2023 Nelson Vieira
%
% @author Nelson Vieira <nelson0.vieira@gmail.com>
% @license CC-BY-4.0 <https://creativecommons.org/licenses/by/4.0/legalcode.txt>
\section{Discussion}

To answer the research question \textbf{RQ3}, the questionnaire makes it clear that there is
a general lack of digital literacy, especially when it comes to IoT.
This still being a new terminology/technology and only quickly expanding
on the last decade, the people that have the most knowledge are the
ones working in areas related to it. This survey also helps to demystify
the privacy paradox.

Participants seemed eager to learn more about IoT, many had no
knowledge of the term, even if some of them knew some devices
that belong to the IoT. Some participants took the time to search
online about terms that appeared on the questionnaire to get some
idea of what they are about, and spend some time discussing them.

The usability tests were very important in improving certain aspects of the
application. As was mentioned in the previous chapter, when the first
usability tests were being conducted, the application was only available
in english, but this soon changed and a portuguese translation was added.
Other improvements added based on participant's feedback was a button
to add an IoT device in the page where is possible to see all available
devices. The icons were also altered, because participants complained that
they were not sure how to read them, even after using for some time.

As concluded in the survey and users do not have a great deal of literacy
regarding IoT, they do have some privacy literacy though, so
to answer the research question \textbf{RQ4}, this application servers to increase users
privacy literacy by giving them tools to know what kind of devices
are around them, what these devices do and give them more information
regarding IoT in general and IoT privacy, so that users may make
well-informed choices.

The application by itself does not provide any formal privacy protection
on IoT devices, but users can use it to better their understanding of IoT
and in some cases make privacy choices regarding a Iot device.

One problem this application faces, and other applications where there is
some kind of user interaction also face, is the fact that some bad
actors will abuse the system by creating many fake accounts, by
data scrapping, by adding bogus data or by replacing existing information
with bogus data.

Some level of theoretical saturation \cite{low2019pragmatic} was reached with the use of the questionnaire
and the usability tests, i.e. it was extracted the most amount of information
possible from the participants on this topic of privacy on IoT systems, since
when doing the usability tests most participants also completed the questionnaire.

This application does not claim to be the best in addressing IoT privacy issues but it improves
on the previous works by not only focusing on privacy choices but also
providing privacy literacy to the end user, which is the ultimate goal
of the application. While existing works are fragmented on their approach,
this application offers a more centralized way by allowing users to complete
all tasks using it. Users can also contribute to the application itself
through various methods, the most straightforward way is using it and
adding new IoT devices to the application's database, they can also leave
feedback for improving the application on distribution platforms. Because
this application is distributed as free and open source software, it
is possible for users to the development of the
application itself by contributing with code or raising questions about
features or bugs, it also
can demystify privacy concerns about the application itself because
anyone can inspect the source code.
