% SPDX-License-Identifier: CC-BY-4.0
%
% Copyright (c) 2023 Nelson Vieira
%
% @author Nelson Vieira <nelson0.vieira@gmail.com>
% @license CC-BY-4.0 <https://creativecommons.org/licenses/by/4.0/legalcode.txt>
\documentclass[conference]{IEEEtran}
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\graphicspath{ {./assets/} }
\usepackage{textcomp}
\usepackage{xcolor,colortbl}
\usepackage{tikz}
\usepackage{adjustbox}
\usepackage[hidelinks]{hyperref}
\hypersetup{
    colorlinks=false,
}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
% \def\myversion{1.0}

\begin{document}

\title{Capacitando os utilizadores para os seus direitos de privacidade na \textit{Internet of Things}\\
}

\author{\IEEEauthorblockN{Nelson Vieira}
\IEEEauthorblockA{\textit{Faculdade de Ciências Exatas e da Engenharia} \\
\textit{Universidade da Madeira}\\
Funchal, Portugal \\
2080511@student.uma.pt}
\and
\IEEEauthorblockN{Mary Barreto (Orientadora)}
\IEEEauthorblockA{\textit{Faculdade de Ciências Exatas e da Engenharia} \\
\textit{Universidade da Madeira}\\
Funchal, Portugal \\
mary.barreto@staff.uma.pt}
}

\maketitle

\begin{abstract}
Os dispositivos da \textit{Internet of Things} (IoT) estão em toda parte, desde o
nascimento da computação ubíqua que a vida quotidiana humana é imaginada
contendo milhões de dispositivos que controlam todos os aspetos de nossas
vidas. Hoje em dia temos carros inteligentes, casas inteligentes, cidades
inteligentes, \textit{wearables} entre outras coisas que usam vários tipos
de dispositivos e vários tipos de redes para comunicarem. Estes dispositivos
criam novas formas de recolher e processar dados pessoais de utilizadores
e não utilizadores. A maioria dos utilizadores finais nem sabem ou têm pouco
controlo sobre as informações que estão a ser recolhidas por estes sistemas.
Este trabalho faz uma abordagem holística deste problema, fazendo primeiro
uma revisão da literatura, depois realizando um questionário para saber mais
sobre o conhecimento geral do público e, finalmente, com base nas informações
recolhidas, é proposto um sistema que fornece aos utilizadores informações
sobre o dispositivos que estão próximos e como proteger os dados que não
desejam compartilhar com estes dispositivos, este sistema é capaz de detetar
que tipo de dispositivos estão próximos, que tipo de dados são recolhidos
por estes dispositivos, mostrar opções de privacidade ao utilizador quando
é possível fazê-lo e o que pode ser feito para evitar que dados indesejados
sejam recolhidos.

% Internet of things (IoT) devices are everywhere, since the birth of ubiquitous
% computing that human every day life is envisioned containing millions of
% devices that control every aspect of our lives. Today we have smart cars,
% smart houses, smart cities, wearables among other things that use various
% types of devices and various types of networks to communicate. These devices
% create new ways of collecting and process personal data from users and
% non-users. Most end users are not even aware or have little control over
% the information that is being collected by these systems. This work takes
% an holistic approach to this problem by first doing a literature review,
% then conducting a survey to learn more about the general knowledge of the
% public, and finally, based on the information gathered, a system is proposed
% that gives users information about the devices that are nearby and how
% to protect the data that they do not want to share with these devices, this
% system is capable of detecting what type of devices are nearby, what kind
% of data is collected by these devices, show privacy choices to the user
% when it is possible to do so and what can be done to protect unwanted data
% from being collected.
\end{abstract}

\begin{IEEEkeywords}
privacidade, Internet of Things, computação ubíqua, assistente de privacidade

% privacy, Internet of Things, ubiquitous computing, privacy assistant
\end{IEEEkeywords}

\section{Introdução}

A privacidade como a conhecemos é um conceito relativamente recente \cite{vincent2016privacy, moore2017privacy},
antes da era digital quase não havia a noção de privacidade para a maioria
das pessoas. Durante muitos séculos, a maioria das pessoas residia em pequenas
comunidades onde estavam continuamente envolvidas na vida umas das outras.
Ainda mais recente é a ideia de que a privacidade é um componente crucial da
segurança pessoal, em contraste com a inegável necessidade da segurança pública,
incluindo a exigência de muros vigiados e portas fechadas. Há muito vista
como um luxo, a privacidade ainda costuma ser vista como um bem de se ter
e não como um requisito essencial, embora seja reconhecida como um direito
humano, conforme consta no artigo 12 da Declaração Universal dos Direitos
Humanos \cite{RooseveltUniversal}: ``Ninguém sofrerá intromissões arbitrárias
na sua vida privada, na sua família, no seu domicílio ou na sua correspondência,
nem ataques à sua honra e reputação. Contra tais intromissões ou ataques
toda a pessoa tem direito a proteção da lei''. A privacidade pode ser definida
\cite{InternationalWhat, SpiekermannEngineering} como o direito de governar
como as informações e dados pessoais são recolhidos, armazenados e usados,
frequentemente envolve o uso de informações confidenciais com cuidado e, como
tal, as organizações devem ser abertas e honestas sobre o tipo de dados que
planeiam recolher, por que precisam deles e onde e com quem planeiam
partilhá-los. Os utilizadores devem ter o direito de controlar como as suas
informações são partilhadas.

% Privacy as we know it is a somewhat recent concept \cite{vincent2016privacy, moore2017privacy},
% before the digital age there was barely any notion of privacy for most
% people. For many centuries most people used to reside in small communities
% where they were continuously involved in one another's lives. Even more
% recent is the idea that privacy is a crucial component of personal security,
% in contrast to the undeniable necessity of public security, including the
% requirement for guarded walls and closed doors. Long seen as a luxury, privacy
% is still usually regarded as a good to have rather than an essential
% requirement, even though it is acknowledged as a human right, as present
% in article 12 of the Universal Declaration of Human Rights \cite{RooseveltUniversal}:
% ``No one shall be subjected to arbitrary interference with his privacy,
% family, home or correspondence, nor to attacks upon his honour and reputation.
% Everyone has the right to the protection of the law against such interference
% or attacks''. Privacy can be defined \cite{InternationalWhat, SpiekermannEngineering}
% as the right to govern how personal information and data is collected, stored,
% and used, it frequently involves handling sensitive information with care,
% and as such, organizations must be open and honest about the kind of data
% they plan to gather, why they need it, and where and with whom they plan
% to share it. Users should have the right to control their shared information.

Esta definição pode causar alguma confusão com a ideia de segurança \cite{HIVDifference}
e embora a privacidade e a segurança estejam interligadas, a segurança envolve
medidas tomadas para salvaguardar os dados de risco, ameaça ou perigo, frequentemente
alude à proteção. É a prática de manter a informação pessoal dos utilizadores
e dados seguros e prevenindo o acesso não autorizado aos mesmos. O principal
contraste entre privacidade e segurança é que o primeiro trata de informações
pessoais dos indivíduos e como estes querem que os seus dados sejam utilizados
e mantidos, enquanto este último trata da sua proteção contra possíveis ameaças.
A segurança pode existem sem privacidade, mas o oposto não é verdade. Para
a gestão de dados sensíveis e pessoais, privacidade e segurança informática
são igualmente cruciais. Os utilizadores devem estar cientes dos procedimentos
internos relativos à recolha, processamento, retenção, e partilha de informação
pessoal.

% This definition can cause some confusion with the idea of security \cite{HIVDifference}
% and although privacy and security are interconnected, security involves
% measures taken to safeguard data from risk, threat or danger, it frequently
% alludes to safety. It is the practice of keeping users' personal information
% and data safe and preventing unauthorized access to it. The primary contrast
% between privacy and security is that the former deals with personal information
% to individuals and how they want their data used and maintained, whilst
% the latter deals with its protection from possible threats. Security can
% exist without privacy, but the opposite is not true. For managing sensitive
% and personal data, privacy and computer security are equally crucial. Users
% should be aware of the internal procedures regarding the collection, processing,
% retention, and sharing of personal information.

As preocupações com a privacidade digital têm vindo a crescer \cite{emami2019exploring, park2022personal, zhang2022peer}
nos últimos anos, especialmente depois dos ataques informáticos do grupo
descentralizado de hackers \textit{Anonymous}, \textit{WikiLeaks} e da divulgação
de documentos ultra secretos da Agência de Segurança Nacional dos Estados
Unidos por Snowden. Estas preocupações podem ser notadas com o aumento da
literatura escrita sobre o assunto, ao procurar termos como ``privacy'',
``online privacy'', ``digital privacy'' no Google Scholar, ACM Digital Library
ou Science Direct, pode ser visto que, nos últimos 5 anos, retornam cerca
de 5 000 000, 650 000 e 8 0000 documentos, respetivamente, incluindo artigos,
livros, artigos de conferência, etc.

% Concerns about digital privacy have been growing \cite{emami2019exploring, park2022personal, zhang2022peer}
% in the last few years, especially after the Anonymous decentralized hacker
% group cyber attacks, WikiLeaks and Snowden's leaked top secret documents
% from United State's National Security Agency. These concerns can be noted
% with the increase of written literature on the subject, when searching for
% terms like ``privacy'', ``online privacy'', ``digital privacy'' in Google
% Scholar, ACM Digital Library or Science Direct it can be seen that, in the
% last 5 years, it returns about 5000000, 650000 and 80000 documents respectively,
% including articles, books, conference papers etc.

A maioria da investigação tem-se concentrado na \textit{web}, enquanto a
privacidade nos sistemas IoT não tem sido tão explorada. Uma vez que os
dispositivos IoT estão a tornar-se mais predominantes, surgem novos métodos
de comunicação, recolha, e análise de dados. Porque já existe uma quantidade
substancial de investigação centrada em privacidade na \textit{web} em vez de
privacidade na IoT, é um terreno muito mais fértil para explorar a questão
da privacidade no contexto da IoT.

% Most research has focused on the web, while privacy in IoT systems has not
% been explored as much. Because IoT devices are becoming more prevalent,
% new methods of communicating, gathering, and analysing data emerge.
% Because there is already a substantial quantity of research focusing on
% web privacy rather than IoT privacy, it is a lot more fertile ground to
% explore the issue of privacy in the context of the IoT.

\textit{Internet of Things} é um termo que surgiu pela primeira vez nos anos
90, e pode estar ligado ao artigo de Mark Weiser sobre a computação ubíqua
\cite{weiser1991computer} e o crescimento de dispositivos de todos os tamanhos
que comunicam entre si para realizar várias tarefas, tornando o sonho de
Weiser uma realidade. A primeira utilização do termo \textit{Internet of
Things} foi em 1999 pelo pioneiro da tecnologia britânica Kevin Ashton \cite{KevinThat},
director executivo do Centro de Auto-ID no Instituto de Tecnologia de Massachusetts,
para descrever um sistema em que os itens podem ser ligados à internet por
sensores. Ele inventou a frase enquanto fazia uma apresentação para a Procter
\& Gamble para realçar o valor de ligar etiquetas de Identificação por Radiofrequência
(RFID) utilizadas em cadeias de fornecimento corporativas ligadas à internet,
a fim de contar e rastrear bens sem a necessidade de assistência humana.
Estes dispositivos são utilizados em várias aplicações, começando em casa
\cite{marikyan2019systematic} com termostatos, frigoríficos, micro-ondas,
etc., passando para carros inteligentes \cite{arena2020overview}, o sistema
educacional \cite{al2020survey}, as nossas roupas e os nossos relógios \cite{niknejad2020comprehensive}
e mesmo para o espaço exterior \cite{AkyildizInternet}. Os recursos da IoT
podem incluir equipamento IoT (como assistentes domésticos inteligentes e
veículos autónomos), serviços IoT (como serviços de análise de vídeo ligados
a câmaras inteligentes e sistemas de localização de posição interior), ou
aplicações IoT (como aplicações remotas de televisão inteligente) que localizam
e utilizam informação sobre nós. A \textit{Internet of Things} é agora amplamente
utilizada para descrever situações em que uma gama de objetos, dispositivos,
sensores, e itens comuns estão ligados à internet e têm capacidades computacionais.

% \textit{Internet of Things} is a term that first appeared in the 1990s,
% and it may be linked to Mark Weiser's paper on ubiquitous computing \cite{weiser1991computer}
% and the growth of devices of all sizes that communicate with one another
% to do various tasks, making Weiser's dream a reality. The first use of the
% term \textit{Internet of Things} was in 1999 by British technology pioneer
% Kevin Ashton \cite{KevinThat}, executive director of the Auto-ID Center
% at Massachusetts Institute of Technology (MIT), to describe a system in
% which items may be connected to the internet by sensors. He came up with
% the phrase while giving a presentation for Procter \& Gamble to highlight
% the value of linking Radio-Frequency Identification (RFID) tags used in
% corporate supply chains to the internet in order to count and track goods
% without the need for human assistance. These devices are used in various
% applications, starting at home \cite{marikyan2019systematic} with thermostats,
% fridges, microwaves, etc, moving on to smart cars \cite{arena2020overview},
% the educational system \cite{al2020survey}, our clothes and our watches \cite{niknejad2020comprehensive}
% and even into outer space \cite{AkyildizInternet}. IoT resources may include
% IoT equipment (like smart home assistants and autonomous vehicles), IoT
% services (like video analytics services linked to smart cameras and indoor
% position tracking systems), or IoT apps (like smart TV remote apps) that
% track and use information about us. Internet of Things is now widely used
% to describe situations in which a range of objects, gadgets, sensors, and
% ordinary items are connected to the internet and have computational capabilities.

A ideia de utilizar computadores e redes para monitorizar e gerir dispositivos
não é nova, apesar do termo \textit{Internet of Things} ser relativamente
recente. As melhorias da tecnologia sem fios nos anos 90 permitiram a adoção
generalizada de soluções máquina-a-máquina (M2M) corporativas e industriais
para monitorização e operação de equipamentos. Muitas das primeiras soluções
M2M, por outro lado, baseavam-se em redes proprietárias ou em normas específicas
da indústria, em vez de normas da internet. Ligar outros dispositivos que
não um computador à internet não é um conceito novo. Uma máquina de Coca-Cola
no Departamento de Informática da Universidade de Carnegie Mellon \cite{EverhartInteresting}
foi o primeiro dispositivo ubíquo a ser ligado à internet. O sistema, criado
em 1982, observou remotamente as luzes de esgotamento de stock nos botões
de pressão da máquina de venda automática e difundiu o estado de cada fila
da máquina de venda automática na rede para que pudesse ser acedido utilizando
o protocolo Name/Finger através de um terminal. Em 1990, uma torradeira que
podia ser ligada e desligada através da internet, criada por John Romkey \cite{RomkeyToast},
foi demonstrada no evento Interop Internet Networking.

% The idea of using computers and networks in order to monitor and manage
% devices is nothing new, despite the term \textit{Internet of Things} being
% relatively recent.
% Wireless technology improvements in the 1990s permitted the widespread
% adoption of corporate and industrial machine-to-machine (M2M) solutions
% for equipment monitoring and operation. Many early M2M solutions, on the
% other hand, relied on proprietary purpose-built networks or industry-specific
% standards rather than internet standards. To connect devices other than
% computers to the internet is not a new concept. A Coke machine at Carnegie
% Mellon University's Computer Science Department \cite{EverhartInteresting}
% was the first ubiquitous device to be linked to the internet. The system,
% which was created in 1982, remotely observed the out of stock lights on
% the pressing buttons of the vending machine and broadcast the state of each
% row of the vending machine on the network so that it could be accessed using
% the Name/Finger protocol through a terminal. In 1990, a toaster that could
% be turned on and off over the internet that was created by John Romkey \cite{RomkeyToast},
% was demonstrated at the Interop Internet Networking show.

A \textit{Internet of Things} pode ser definida como: ``Uma rede aberta e
abrangente de objetos inteligentes que têm a capacidade de se auto-organizarem,
partilhar informação, dados e recursos, reagindo e agindo face a situações
e mudanças no ambiente'' \cite{madakam2015internet}.

% The Internet of Things can be defined as: ``An open and comprehensive network
% of intelligent objects that have the capacity to auto-organize, share information,
% data and resources, reacting and acting in face of situations and changes
% in the environment'' \cite{madakam2015internet}.

IoT é uma das tecnologias de crescimento mais rápido \cite{MohammadState},
prevê-se que cresça para os triliões de dispositivos até 2030 \cite{SarawiInternet},
e com esta expansão surgem novas vulnerabilidades de segurança e perigos
de recolha de dados, a falta de segurança nestes dispositivos torna-os alvos
ideais para violações da privacidade e divulgação inadequada aos clientes
das capacidades dos dispositivos e práticas de dados agrava as questões de
privacidade e segurança.

% IoT is one of the fastest growing technologies \cite{MohammadState}, it
% is predicted that it will grow into the trillions of devices by 2030 \cite{SarawiInternet},
% and with this expansion new security vulnerabilities and data gathering
% dangers appear, the lack of security in these devices makes them ideal targets
% for privacy violations and inadequate customer disclosure of device capabilities
% and data practices aggravates privacy and security issues.

A privacidade nos sistemas IoT não é vista como um factor crucial no desenvolvimento
de aplicações ou sistemas \cite{alhirabi2021security}. Normas específicas
para opções de privacidade foram impostas por normas de privacidade dos dados
incluindo o Regulamento Geral de Proteção de Dados (RGPD) e a Lei de Privacidade
do Consumidor da Califórnia, mas mesmo estes regulamentos têm sido criticados \cite{peloquin2020disruptive, gladis2022weaponizing, gentile2022deficient, green2022flaws, byun2019privacy}.

% Privacy in IoT systems in not seen as a crucial factor in development \cite{alhirabi2021security}.
% Specific standards for privacy options have been imposed by data privacy
% regulations including the General Data Protection Regulation (GDPR) and
% California Consumer Privacy Act (CCPA), but even these regulations have
% been criticized \cite{peloquin2020disruptive, gladis2022weaponizing, gentile2022deficient, green2022flaws, byun2019privacy}.

\section{Estado da Arte}

\par
Esta secção fornece uma visão geral da literatura recente com os temas que
foram consideradas mais relevantes para este trabalho.

% This section provides an overview of the recent literature with the themes
% that were found to be more relevant for this work.

\subsection{Paradoxo da Privacidade}

A utilização de uma variedade de dispositivos digitais tem inúmeras vantagens,
mas eles trazem também consigo a ubiquidade do equipamento de captura de dados,
é compreensível que a maioria dos utilizadores em rede tenha sérias preocupações
sobre a privacidade dos seus dados pessoais. No entanto, as opiniões expressas
estão em total desacordo com a realidade, segundo o relatório sobre o estado
da privacidade de Thomson et al. \cite{DarrenState}, de apenas um em cada
quatro utilizadores europeus lê os termos e condições na sua totalidade antes
de efetuar uma compra \textit{on-line} ou subscrever a um serviço, 59\% admitiram
apenas a digitalização rápida do termos e condições antes de concluir uma
compra, enquanto 14\% admitiu que nunca os leu de todo, 30\% dos inquiridos
até trocariam os seus endereço de \textit{e-mail} para ganhar uma recompensa,
ou entrada numa rifa, enquanto 17\% faria por isso, para obter uma aplicação
e 30\% fá-lo-ia por dinheiro.

% The use of a variety of digital devices have numerous advantages, but they
% also bring with them the ubiquity of data capturing equipment, therefore,
% it is understandable why the majority of online users have serious concerns
% about the privacy of their personal data. However, the opinions expressed
% are starkly at odds with the reality, according to Thomson et al. \cite{DarrenState}
% report on the state of privacy, that just one in four European users read
% the terms and conditions in their entirety prior to making an online purchase
% or subscribing to a service, 59\% admitted to only quickly scanning the
% terms and conditions before completing a purchase, while 14\% admitted to
% never reading them at all, 30\% of the respondents would even swap their
% email address to win a reward, or entry into a raffle, while 17\% would do
% so to get an app and 30\% would do it for money.

Isto é o que se chama um paradoxo de privacidade, tem havido múltiplos artigos
escritos sobre este assunto \cite{solove2021myth, WilliamsPrivacy, lee2021investigating, goad2021privacy, gerber2018explaining},
alguns artigos tentam uma explicação teórica enquanto outros tentam uma explicação
empírica. Tem havido interpretações ou explicações muito diferentes deste
paradoxo, alguns artigos \cite{wilson2012unpacking, warshaw2015can, lee2015privacy}
aplicam o conceito teórico do \textit{homo economicus} \cite{zak2008moral},
que é a representação de pessoas como seres que atuam constantemente de uma
forma lógica e interessada em si próprios, sem se preocuparem com a moralidade
ou a ética, e que o fazem o melhor que podem, no contexto da privacidade.
Diferentes preconceitos cognitivos e heurísticos podem influenciar a forma
como os consumidores tomam decisões, de acordo com vários estudos sobre o
comportamento de escolha do consumidor \cite{acquisti2007can, knijnenburg2013dimensionality, wakefield2013influence, flender2012type}.
De acordo com vários artigos \cite{dienlin2015privacy, baek2014solving}, este
paradoxo pode ser explicado pelo facto de algumas pessoas terem genuinamente
violações de privacidade online e que a maioria das opiniões sobre privacidade
são, portanto com base na heurística ou em contas de segunda mão. O estudo
de Taddicken \cite{taddicken2014privacy} argumenta que a pressão de grupo
é a razão pela qual as pessoas têm este comportamento contraditório, Norberg
et al. \cite{norberg2007privacy} explica este paradoxo ao sugerir que embora
a perceção do risco afete as atitudes e intenções comportamentais relatadas,
a confiança tem um impacto direto no comportamento de privacidade, enquanto
outros \cite{flender2012type, kokolakis2017privacy} confiam na teoria quântica.
Brandimarte et al. \cite{brandimarte2013misplaced} exploraram a ideia de que
quando se trata da sua privacidade de dados, os utilizadores têm uma \textit{ilusão
de controlo}.

% This is what is called a privacy paradox, there have been multiple papers
% written on this subject \cite{solove2021myth, WilliamsPrivacy, lee2021investigating, goad2021privacy, gerber2018explaining},
% some papers attempt a theoretical explanation while others attempt an empirical
% one. There has been very different interpretations or explanations of this
% paradox, a few papers \cite{wilson2012unpacking, warshaw2015can, lee2015privacy}
% apply the theoretical concept of the \textit{homo economicus} \cite{zak2008moral},
% which is the representation of people as beings who constantly act in a
% way that is logical and self-interested, not worrying about morality or
% ethics, and who do so to the best of their ability, to the context of privacy.
% Different cognitive biases and heuristics can influence how consumers make
% decisions, according to several studies on consumer choice behaviour \cite{acquisti2007can, knijnenburg2013dimensionality, wakefield2013influence, flender2012type}.
% According to several articles \cite{dienlin2015privacy, baek2014solving},
% this paradox might be explained by the fact that some people have genuinely
% experienced online privacy assaults and that most privacy views are therefore
% based on heuristics or secondhand accounts. Taddicken's study \cite{taddicken2014privacy}
% argues that peer pressure is the reason people have this contradictory behaviour,
% Norberg et al. \cite{norberg2007privacy} explains this paradox by suggesting
% that while perceived risk affects reported attitudes and behavioural intentions,
% trust has a direct impact on privacy behaviour, while others \cite{flender2012type, kokolakis2017privacy}
% rely on quantum theory. Brandimarte et al. \cite{brandimarte2013misplaced}
% have explored the idea that when it comes to their data privacy, users have
% an \textit{illusion of control}.

Este paradoxo foi comprovado por uma série de estudos empíricos \cite{dienlin2015privacy, xie2019consumers, SCHWAIG20131, sannon2018privacy},
as práticas de privacidade online são baseadas em mentalidades de privacidade
separadas e assim não são intrinsecamente paradoxais.

% This paradox has been proven to be vitiated by a number of empirical studies \cite{dienlin2015privacy, xie2019consumers, SCHWAIG20131, sannon2018privacy},
% online privacy practices are founded on separate privacy mindsets and so
% they are not inherently paradoxical.

\subsection{Privacidade em IoT: Abordagens}

Houve uma série de revisões sistemáticas da literatura (SLR) \cite{Gupta2022Privacy, Kuhtreiber2022survey, sicari2015security, LinSurvey}
e \textit{systematic mapping reviews} \cite{porras2018security, ahmed2019aspects}
feitas para estudar as questões de privacidade e segurança na IoT.

% There have been a number of systematic literature reviews (SLR) \cite{Gupta2022Privacy, Kuhtreiber2022survey, sicari2015security, LinSurvey}
% and systematic mapping reviews \cite{porras2018security, ahmed2019aspects}
% done to study privacy and security issues in IoT.

No SLR de Gupta e Ghanavati \cite{Gupta2022Privacy}, os autores revêm os
artigos com metodologias e técnicas que identificam riscos de privacidade
ou que notificam os utilizadores sobre estes riscos. Eles dividem a literatura
nas seguintes categorias: `Modelação Ontológica e Abordagens Semânticas',
`Abordagens de Dados', `Abordagens Baseadas na Análise do Código Fonte',
`Estudos do Utilizador e Abordagens Baseadas em Inquéritos', `Abordagens
Baseadas em \textit{Blockchain}' e `Abordagens Baseadas em Estruturas e Arquiteturas'.
Em seguida, examinam a literatura actual sobre estes três pré-requisitos.
Os resultados mostram que: a maioria das obras concentra-se em dispositivos
IoT únicos quando se trata de ameaças à privacidade; ao analisar as questões
de privacidade, fatores-chave de privacidade tais como redução e agregação
de dados são ignorados; os estudos existentes ignoraram a sensibilidade das
informações obtidas; os estudos mais úteis não incluíram um leque diversificado
de utilizadores ao avaliar os problemas de privacidade; não foi feito nenhum
trabalho para descobrir as dificuldades de conformidade entre uma aplicação
IoT e diferentes regras de privacidade; e de acordo com os dados atuais,
fornecer aos clientes notificações de privacidade em tempo real não é uma
prioridade elevada. No entanto, este SLR tem as seguintes limitações: os autores
apenas escolheram artigos e não teses ou livros e a partir dos trabalhos selecionados,
só foram considerados os escritos em inglês.

% In Gupta and Ghanavati's \cite{Gupta2022Privacy} SLR, the authors review
% papers with methodologies and techniques that identify privacy risks or
% notify users about these risks. They divide the literature into the following
% categories: `Ontological Modelling and Semantic-based Approaches', `Data-Driven
% Approaches', `Source Code Analysis-based Approaches', `User Studies and
% Survey-based Approaches', `Blockchain-based Approaches' and `Architectural
% and Framework-based Approaches'. They then examine current literature on
% these three prerequisites. The findings show that: most works concentrate
% on single IoT devices when addressing privacy threats; When analysing privacy
% issues, key privacy factors such as data reduction and data aggregation
% are overlooked; existing studies ignored the sensitivity of the obtained
% information; most useful studies did not include a diverse range of users
% when assessing privacy problems; no work has been done to discover compliance
% difficulties between an IoT application and different privacy rules; and
% current research does not place a premium on providing consumers with real-time
% privacy notices. However, this SLR has the following limitations: the authors
% only chose articles and not thesis or books and from the selected papers,
% only the ones written in english were considered.

Kühtreiber et al. \cite{Kuhtreiber2022survey} avaliam as \textit{frameworks}
e ferramentas estabelecidas para programadores, especificamente no caso da
IoT, e acham que as soluções atuais são difíceis de utilizar, apenas com
sucesso em cenários limitados e insuficiente para lidar com os problemas
de privacidade inerentes ao desenvolvimento de IoT. Este estudo carece de
uma revisão exaustiva das lacunas escolhidas literatura, juntamente com questões
de investigação que estabeleçam o significado dos artigos escolhidos.

% Kühtreiber et al. \cite{Kuhtreiber2022survey} evaluate the frameworks and
% tools established for developers, specifically in the case of IoT, and find
% that current solutions are difficult to use, only successful in limited
% scenarios, and insufficient to handle the privacy problems inherent in IoT
% development. This study lacks a comprehensive gap review of the chosen
% literature, along with research questions establishing the significance
% of the articles chosen.

Sicari et al. \cite{sicari2015security} examinam a investigação em curso
e atividades em curso que se concentram em soluções de privacidade e segurança
da IoT. Os autores começar por descrever os requisitos de privacidade e
segurança da IoT, tais como controlo de acesso, confidencialidade, e autenticação.
Os autores então realizaram um estudo de literatura relacionado com estas
três necessidades. Os autores chegaram à conclusão de que as questões de
privacidade da IoT foram apenas parcialmente examinadas e que é necessária
mais atenção. O estudo, no entanto, tem falhas: a análise prévia da investigação
centra-se principalmente nas necessidades de segurança, enquanto ignorando
considerações de privacidade; os autores não realizam uma análise exaustiva
das lacunas nas publicações examinadas; e não fornecem um resumo abrangente
de futuros tópicos de investigação no domínio da privacidade na IoT que requerem
mais atenção.

% Sicari et al. \cite{sicari2015security} examine current research and ongoing
% activities that focus on IoT privacy and security solutions. The authors
% start by describing the requirements for IoT privacy and security, such
% as access control, confidentiality, and authentication. The authors then
% conduct a literature study in connection to these three needs. The authors
% came to the conclusion that IoT privacy issues have only been partially
% examined and that further attention is required. The study, however, has
% flaws: the prior research analysis focuses primarily on security needs while
% ignoring privacy considerations; the authors do not conduct a thorough gap
% analysis on the publications examined; and they do not provide a comprehensive
% summary of future research topics in the field of IoT privacy that require
% more attention.

Lin et al. \cite{LinSurvey} realizam uma revisão da literatura para identificar
vulnerabilidades de segurança e privacidade nas três camadas da arquitetura
da IoT: rede, perceção, e aplicação. Os autores descrevem as seis primeiras
propriedades fundamentais de segurança para estes níveis como confidencialidade,
integridade, disponibilidade, identificação e autenticação, privacidade,
e confiança. Em seguida, os autores analisam uma variedade de ameaças à segurança
para cada uma das três fases. Os autores concluem dando um resumo sucinto
de muitas técnicas de preservação da privacidade dos dados, incluindo as
fases de recolha de dados, agregação e análise dos dados. Os autores concentram-se,
contudo, em grande medida nas componentes de segurança da IoT e, como já
foi dito, consideram a privacidade como um dos aspetos de segurança mais
cruciais, em vez de encararem a privacidade como uma preocupação distinta.
Além disso, a investigação não realiza uma análise exaustiva de lacunas para
descobrir os pontos fracos dos esforços anteriores.

% Lin et al. \cite{LinSurvey} undertake a literature review to identify security
% and privacy vulnerabilities in the three IoT architecture layers: network,
% perception, and application. The authors describe the first six fundamental
% security properties for these tiers as confidentiality, integrity, availability,
% identification and authentication, privacy, and trust. Then, the authors
% look at a variety of security threats for each of the three stages. The
% authors wrap up by giving a succinct summary of many privacy-preserving
% data techniques, including the stages of data collection, data aggregation,
% and data analysis. The authors do, however, largely focus on the IoT's security
% components and, as was already said, consider privacy to be one of the most
% crucial security aspects, rather than viewing privacy as a distinct concern.
% Furthermore, the research does not conduct a thorough gap analysis to discover
% the weaknesses of prior efforts.

Com base na análise da literatura de Ziegeldorf \cite{ziegeldorf2014privacy},
seguem-se as preocupações mais proeminentes em matéria de privacidade na
IoT:

% Based on Ziegeldorf's \cite{ziegeldorf2014privacy} analysis of the literature,
% the following are the most prominent privacy concerns in IoT:

\begin{enumerate}
    \item
    A preocupação mais proeminente é a \textit{identificação}, que liga
    um identificador, tal como um nome e localização, com a identidade de
    um indivíduo, isto também permite e agrava outras ameaças;

    % The most prominent concern is \textit{identification}, which binds an
    % identifier, such as a name and location, with an individual's identity,
    % this also enables and aggravates other threats;
    \item
    \textit{Localização e rastreio} é a ameaça de detetar a localização
    de um indivíduo através de numerosas técnicas, tais como GPS, tráfego
    na internet, ou localização do smartphone. Esta ameaça requer \textit{identificação}
    de algum tipo;

    % \textit{Localization and tracking} is the threat of detecting an individual's
    % locations through numerous techniques, such as GPS, internet traffic,
    % or smartphone location. This threat requires \textit{identification}
    % of some kind;
    \item
    No comércio eletrónico, o \textit{profiling} é frequentemente utilizado
    para personalização. As organizações recolhem informação sobre indivíduos
    a fim de deduzir os seus interesses através da associação com outros
    perfis e fontes de dados.

    % In e-commerce, \textit{profiling} is often used for personalization.
    % Organizations collect information about individuals in order to deduce
    % their interests via association with other profiles and data sources.
    \item
    \textit{Interação e apresentação} faz alusão à partilha de informação
    privada com um público não intencional enquanto o faz através de um
    meio público. As aplicações IoT necessitam frequentemente de uma interação
    extensa com o utilizador, espera-se que os utilizadores destes sistemas
    obtenham informações através de dispositivos inteligentes no seu ambiente
    imediato e que os utilizadores interagem com sistemas de formas criativas
    e naturais. No entanto, muitos desses modos de comunicação e apresentação
    já estão disponíveis para o público em geral, tornando-os aparentes
    a qualquer pessoa por perto. Quando a informação pessoal é transferida
    entre um sistema e o seu utilizador, a privacidade é violada.

    % \textit{Interaction and presentation} allude to the sharing of private
    % information with an unintended audience while doing so through a public
    % medium. IoT applications often need extensive user interaction, it is
    % expected that users of these systems will obtain information via smart
    % devices in their immediate surroundings and that users will interface
    % with systems in creative, natural ways. However, many of those modes
    % of communication and presentation are already available to the broader
    % public, making them apparent to anybody around. When personal information
    % is transferred between a system and its user, privacy is breached.
    \item
    \textit{Transições do ciclo de vida} ocorrem quando um dispositivo IoT
    é vendido, utilizado pelo seu proprietário e eventualmente eliminado.
    Pode haver uma expectativa que o objeto apaga todas as informações,
    mas dispositivos inteligentes frequentemente mantêm grandes volumes de
    dados sobre o seu próprio passado ao longo de toda a sua existência.
    Isto pode conter imagens e vídeos pessoais, que nem sempre são apagados
    após a transferência de propriedade.

    % \textit{Lifecycle transitions} occur when an IoT device is sold, utilized
    % by its owner and eventually disposed of. There may be an expectation
    % that the object deletes all information, yet smart devices frequently
    % keep massive volumes of data about their own past throughout their entire
    % existence. This might contain personal images and videos, which are
    % not always erased following ownership transfer.
    \item
    \textit{Ataques ao inventário} envolvem a entrada não autorizada e a
    aquisição de informação sobre a presença e as características das coisas
    pessoais. Os utilizadores maliciosos podem utilizar dados de inventário
    para traçar o perfil da propriedade e assaltar.

    % \textit{Inventory attacks} involve unauthorized entry and the acquisition
    % of information about the presence and characteristics of personal things.
    % Malicious users might use inventory data to profile the property and
    % break in.
    \item
    A \textit{interligação} é o processo de ligar sistemas díspares, quando
    os sistemas estão a ligar diferentes fontes de dados, existe um maior
    perigo de acesso não autorizado e de fuga de dados.

    % \textit{Linkage} is the process of connecting disparate systems, when
    % systems are connecting different data sources, there is a higher danger
    % of unauthorized access and data leakage.
\end{enumerate}

Outro conceito que vale a pena analisar é a privacidade diferencial que
se relaciona mais com o inquérito que será conduzido, mas também com a recolha
geral e análise de dados dos utilizadores por aplicações e sistemas.

% Another concept worth analysing is differential privacy which relates more
% closely to the survey that will be conducted but also to the general
% collection and analysis of user data by applications and systems.

\subsection{Privacidade Diferencial}

A noção de privacidade diferencial, de acordo com Michael Kearns \cite{kearns2019ethical},
baseia-se em três princípios importantes. O primeiro é que ``a privacidade
diferencial exige que a adição ou remoção do registo de dados de um único
indivíduo não alterar a probabilidade de qualquer resultado por muito''.
O segundo princípio sendo que ``nenhum observador exterior pode aprender
muito sobre qualquer indivíduo por causa dos dados específicos dessa pessoa''.
O terceiro princípio importante sendo que ``para cada indivíduo no conjunto
de dados, e para qualquer observador não independentemente das suas crenças
iniciais sobre o mundo, depois de observarem o resultado de um cálculo diferencialmente
privado, a sua crença posterior sobre qualquer coisa está perto do que teria
sido se tivessem observado a saída do mesmo cálculo executado sem os dados
do indivíduo''.

% The notion of differential privacy, according to Michael Kearns \cite{kearns2019ethical},
% is based on three important principles. The first being that ``differential
% privacy requires that adding or removing the data record of a single individual
% not change the probability of any outcome by much''. The second principle
% being that ``no outside observer can learn very much about any individual
% because of that person's specific data''. The third important principle
% being that ``for every individual in the dataset, and for any observer no
% matter what their initial beliefs about the world were, after observing
% the output of a differentially private computation, their posterior belief
% about anything is close to what it would have been had they observed the
% output of the same computation run without the individual's data''.

A privacidade diferencial tem o potencial de aumentar significativamente a
proteção da privacidade individual, ao adicionar propositadamente ruído a
um conjunto de dados, dá negabilidade plausível a qualquer indivíduo que possa
ter tido os seus dados explorados enquanto ainda é capaz de calcular estatísticas
com uma precisão relativamente elevada. Embora os algoritmos que lidam com
noções de justiça, ética e privacidade são difíceis de implementar devido à
subjetividade destes conceitos, e os algoritmos de privacidade diferencial
não são diferentes, ainda podem ajudar em no que diz respeito à abordagem
dos dilemas morais inerentes à tecnologia.

% Differential privacy has the potential to significantly increase individual
% privacy protection, by purposefully adding noise into a dataset, it gives
% plausible deniability to any individual who may have had their data exploited
% while still being able to calculate statistics with relatively high precision.
% Although algorithms that deal with notions of fairness, ethics, and privacy
% are hard to implement because of the subjectivity of these concepts, and
% differential privacy algorithms are no different, they can still help in
% regards to addressing technology's inherent moral quandaries.

Existem outros algoritmos que visam preservar a privacidade da mesma forma
que a privacidade diferencial, tais como o algoritmo de \textit{box blurring}
da Google \cite{FromeLarge} que é utilizado na vista de rua do Google Maps,
o Visor da Microsoft \cite{poddar2020visor} que é uma ferramenta de análise
de vídeo como serviço e o sistema de aprendizagem profunda colaborativa de
Shokri e Shmatikov \cite{ShokriPrivacy}, no entanto, em geral, estes algoritmos
lutam com elevados custos computacionais, ataques internos, ou privacidade
não comprovável.

% There exist other algorithms that aim to preserve privacy in the same way
% as differential privacy such as Google's box blurring algorithm \cite{FromeLarge}
% that is used in the Google Map's street view, Microsoft's Visor \cite{poddar2020visor}
% which is a video-analytics-as-a-service tool and Shokri and Shmatikov's
% \cite{ShokriPrivacy} system for collaborative deep learning, however, in
% general, these algorithms struggle with high computational cost, internal
% attacks, or non-provable privacy.

Zhao et al. \cite{ZhaoSurvey} conduzem um SLR sobre a privacidade diferencial
para dados não estruturados. Os autores apresentam métodos de privacidade
diferencial para conteúdos sensíveis em dados de imagem, áudio, vídeo, e texto.
Comparam os vários métodos e efetuam análises de utilidade para cada método,
destacando os benefícios e desvantagens de cada um, a perda de utilidade é
medida em avaliações experimentais entre os dados reais e a sua variante ofuscada.
Chegam à conclusão de que a privacidade diferencial, bem como as suas variações,
dão proteção rigorosa à privacidade de dados não estruturados contra atacantes
com conhecimentos de base imprevisíveis. Sugerem também potenciais temas de
estudo futuros que ainda têm de ser investigados.

% Zhao et al. \cite{ZhaoSurvey} conduct a SLR on differential privacy for
% unstructured data. The authors present differential privacy methods for
% sensitive content in image, audio, video, and text data. They compare the
% various methods and perform utility analyses for each method, highlighting
% the benefits and drawbacks of each, the utility loss is measured in experimental
% evaluations between the actual data and its obfuscated variant. They come
% to the conclusion that differential privacy as well as its variations give
% stringent privacy protections for unstructured data against attackers with
% unpredictable background knowledge. They also suggest potential future study
% subjects that have yet to be investigated.

\subsection{Soluções Propostas}

\par
Esta secção enumera sete soluções que emergiram da revisão de literatura
estruturada para melhorar o fosso entre os conceitos de privacidade e segurança
entre sistemas e utilizadores.

% This section list seven solutions that emerged from the structured
% literature review to improve the gap between privacy and security concepts
% among systems and users.

\subsubsection{Criando Novas Formas de Conscienlização para o Utilizador}

Tem sido feito algum trabalho para determinar o conhecimento dos utilizadores
sobre as suas ações online relativamente à sua privacidade. Skirpan et al.
\cite{SkirpanPrivacy} construíram uma experiência de teatro interativo, esta
foi criada para tentar provar que uma experiência simulada com um problema
de privacidade credível pode encorajar as pessoas a tomar medidas antes de
se depararem realmente com uma catástrofe. O enredo da peça consiste numa
companhia tecnológica inexperiente que revelou a sua revolucionária tecnologia
de IA enquanto lidava com um denunciante da empresa e um hack inoportuno de
dia zero no seu sistema. O público é capaz de interagir com os atores e influenciar
a forma como a história se desenrola. O público e os atores tiveram a oportunidade
de experimentar papéis, comportamentos e opiniões a que normalmente não teriam
acesso na vida quotidiana. Os autores tiveram entrevistas e inquéritos feitos
após as peças com membros do público, no entanto, apenas fizeram entrevistas
a meio da produção e apenas a uma pequena fração do público participou efetivamente
nesta recolha de dados, observaram também que, após contactarem as pessoas
meses após as entrevistas, não mudaram realmente o seu comportamento relativamente
aos seus direitos de privacidade.

% There has been some work done to determine the users awareness of their
% actions online regarding their privacy. Skirpan et al. \cite{SkirpanPrivacy}
% built an interactive theatre experience, this was created to try to prove
% that a simulated experience with a credible privacy problem may encourage
% people to take action before actually encountering a catastrophe. The plot
% of the play consist in a fledgling tech company that unveiled its revolutionary
% AI technology while dealing with a company whistleblower and an untimely
% zero-day hack on their system. The public is able to interact with the actors
% and influence how the story plays out. Audiences and actors were given the
% chance to try on roles, behaviors, and opinions that they would not normally
% have access to in ordinary life. The authors had interviews and surveys
% done after the plays with audience members however they only did interviews
% halfway through production and only a small fraction of the audience actually
% participated in this data collection, they also noted that after contacting
% people months after the interviews that they did not really changed their
% behaviour regarding their privacy rights.

\subsubsection{Legislação}

Alguns documentos procuram melhorar a legislação \cite{WEBER2015618, FabianoInternet}
porque, caso contrário, nas suas opiniões, os direitos de privacidade não serão
respeitados se não forem legalmente exigíveis, defendem que, sem o acordo expresso
do indivíduo em causa, as informações privadas obtidas por dispositivos IoT
não devem ser retidas ou processadas sob qualquer forma, e devem ser adotados
os procedimentos necessários para garantir que os dados recolhidos não sejam
os de um indivíduo não relacionado. Mas melhores leis de proteção para o utilizador
também criariam oposição da maioria das empresas que pretendem extrair o máximo
de dados privados dos seus utilizadores sem quaisquer restrições, a fim de
aumentar as suas margens de lucro.

% Some papers seek to improve legislation \cite{WEBER2015618, FabianoInternet}
% because otherwise, in their view, privacy rights won't be respected if they
% are not enforceable legally, they defend that without the express agreement
% of the individual concerned, private information obtained by IoT devices
% must not be retained or processed in any form, and necessary procedures
% must be taken to guarantee that the data collected is not that of an unrelated
% individual. But better protection laws for the user would also create opposition
% from most companies that want to extract as much private data from their
% users without (m)any restrictions in order to increase their profit margins.

\subsubsection{Privacidade Através da Segurança}

Sun et al. \cite{SunSecure} concebem uma estratégia de comunicação leve para
um sistema de controlo remoto, utilizando dois tipos de espaços virtuais para
alcançar o objetivo de anúncio de identidade e troca de dados. Construíram
um sistema protótipo do esquema e testaram-no na Freenet, demonstrando que
o método pode resistir eficazmente à influência da análise de fluxo no anonimato
da comunicação, preservando ao mesmo tempo a segurança dos dados de comunicação.

% Sun et al. \cite{SunSecure} design a lightweight communication strategy
% for a remote-control system, employing two types of Virtual-Spaces to achieve
% the aim of identity announcement and data exchange. They constructed a prototype
% system of the scheme and tested it on the Freenet, demonstrating that the
% method can effectively resist the influence of flow analysis on communication
% anonymity while preserving communication data security.

\subsubsection{Propostas de Arquiteturas / Frameworks}

Antunes et al. \cite{AntunesFederated} fazem um SLR sobre aprendizagem federada
na área dos cuidados de saúde e fazem uma proposta de arquitetura. A técnica
conhecida como aprendizagem federada permite a formação distribuída de modelos
de aprendizagem de máquinas utilizando conjuntos de dados hospedados remotamente
sem a necessidade de amplificação de dados. O objetivo fundamental da arquitetura
proposta é permitir que as instituições de saúde que têm acesso a informação
médica sensível a utilizem na análise de dados distribuídos e na investigação
da aprendizagem automática, assegurando ao mesmo tempo a confidencialidade dos
pacientes. Uma vez que a informação transmitida entre instituições necessita
de garantias de confidencialidade para os parâmetros do modelo de aprendizagem
e resultados de análise, a arquitetura pode adotar uma série de formas baseadas
num paradigma de segurança de confiança zero \cite{ChenSecurity}. Além disso,
as instituições desenvolvem um sistema de verificação de algoritmos de aprendizagem
que pode armazenar e disseminar manifestos, bem como envolver-se em procedimentos
analíticos distribuídos que necessitam do acordo unânime de todos os participantes.
Este estudo demonstra também que a literatura anterior implica que a encriptação
homomórfica e a privacidade diferencial são abordagens eficazes para prevenir
violações de dados sem incorrer em custos de computação proibitivamente elevados.

% Antunes et al. \cite{AntunesFederated} do a SLR on federated learning in
% the area of healthcare and make an architecture proposal. The technique
% known as federated learning allows for the distributed training of machine
% learning models using remotely hosted datasets without the requirement for
% data amplification. The fundamental goal of the proposed architecture is
% to allow healthcare institutions that have access to sensitive medical information
% to use it in distributed data analysis and machine learning research while
% ensuring patient confidentiality. Because information transmitted among
% institutions need confidentiality guarantees for learning model parameters
% and analysis results, the architecture can adopt a number of ways based on
% a zero-trust security paradigm \cite{ChenSecurity}. Furthermore, the institutions
% develop a learning algorithm verification system that can store and disseminate
% manifestos, as well as engage in distributed analytic procedures that need
% unanimous agreement from all participants. This study also demonstrates
% that previous literature implies that homomorphic encryption and differential
% privacy are effective approaches for preventing data breaches without incurring
% prohibitively high computing costs.

Opara et al. \cite{opara2022framework} apresentam um sistema para detetar
possíveis problemas com regulamentos de privacidade ou segurança nas fases
iniciais de desenvolvimento, esta abordagem destina-se aos programadores. O
documento propõe uma ontologia de domínio específico para a modelagem de políticas
de segurança e privacidade da IoT, uma notação para representar e validar políticas
de segurança e privacidade da IoT, um conjunto de diretrizes e regras para
detetar erros de políticas da IoT, e uma ferramenta para modelar e capturar
visualmente políticas de segurança da IoT e descobrir problemas de políticas.
Embora a framework apresentada seja teoricamente promissora, ainda não foi
testada num ambiente real, pelo que a sua eficácia ainda não pode ser medida.
Os autores também não comparam a sua proposta com outras já disponíveis.

% Opara et al. \cite{opara2022framework} present a system for spotting possible
% problems with privacy or security regulations in the early stages of development,
% this approach is intended at developers. The paper proposes a domain-specific
% ontology for modelling IoT security and privacy policies, a notation for
% representing and validating IoT security and privacy policies, a set of
% guidelines and rules for detecting IoT policy errors, and a tool for visually
% modelling and capturing IoT security policies and discovering policy problems.
% Although the framework that is presented is theoretically promising it has
% not been tested in a real environment so the effectiveness can't yet be
% measured. The authors also do not compare their proposal with others already
% available.

\subsubsection{Blockchain}

Blockchain é uma opção para garantir a privacidade em IoT devido a provas de
conhecimento zero, assinaturas de anéis e \textit{mixing} \cite{PrivacyblockchainWikipedia}.

% Blockchain is an option to guarantee privacy in IoT because of zero-knowledge
% proofs, ring signatures and mixing \cite{PrivacyblockchainWikipedia}.

Yu et al. \cite{yu2018blockchain} mostra várias implementações de blockchain
que proporcionam privacidade através da segurança, com base em diferentes
categorias como integridade de dados, partilha de dados e autenticação e controlo
de acesso. Os autores utilizam a privacidade como um proxy para a segurança,
também não discutem os pontos fracos e fortes de cada implementação ou fazem
qualquer comparação, também não proporcionam quaisquer outras questões de investigação.

% Yu et al. \cite{yu2018blockchain} shows various implementations of blockchain
% that provide privacy through security, based on different categories like
% data integrity, data sharing and authentication and access control. The
% authors use privacy as a proxy for security, they also do not discuss the
% weak and strong points of each implementation or make any comparison, they
% also do not provide further research questions.

Ali et al. \cite{AliIoT} sugerem um stack de software que combina a partilha
de ficheiros \textit{peer-to-peer} com contratos inteligentes de blockchain
para oferecer aos utilizadores de IoT o controlo sobre os seus dados e eliminar
a necessidade de uma gestão centralizada de dados de IoT. Os contratos inteligentes
de blockchain são utilizados na arquitetura proposta do `consórcio modular'
para regular o acesso, estabelecendo simultaneamente a responsabilidade tanto
dos proprietários dos dados como de outras partes a que os utilizadores concedem
acesso.

% Ali et al. \cite{AliIoT} suggest a software stack that combines peer-to-peer
% file sharing with blockchain smart contracts to offer IoT users control
% over their data and do away with the necessity for centralized IoT data
% management. Blockchain smart contracts are used in the proposed `modular
% consortium' architecture to regulate access while establishing responsibility
% for both data owners and other parties that users grant access to.

\subsubsection{Outras Propostas}

Zhu et al. \cite{ZhuIntegrating} apresentam um sistema de sensor híbrido que
salvaguarda a privacidade, ao mesmo tempo que monitoriza a disponibilidade de
estacionamento. Os autores fundiram a deteção IoT com \textit{crowdsensing}
e melhoraram-na com métodos de preservação da privacidade. Os autores empregaram
filtros físicos nebulosos para mascarar os sensores IoT na deteção de IoT,
e uma técnica criptográfica baseada em compromissos criptográficos, provas
de conhecimento zero, e credenciais anónimas em \textit{crowdsensing}. Além
disso, utilizaram o \textit{crowdsourcing} para criar um modelo de aprendizagem
de máquinas para reconhecimento de estacionamento na presença de filtros de
nevoeiro. O seu papel incluía protótipos de prova de conceito como um sistema
Raspberry Pi e uma aplicação móvel, bem como um estudo de avaliação do modelo
de aprendizagem de máquinas e os efeitos do \textit{crowdsourcing}.

% Zhu et al. \cite{ZhuIntegrating} present a hybrid sensor system that safeguards
% privacy while also monitoring parking availability. The authors merged IoT
% sensing with crowdsensing and enhanced it with privacy-preserving methods.
% The authors employed physical hazy filters to mask IoT sensors in IoT sensing,
% and a cryptographic technique based on cryptographic commitments, zero-knowledge
% proofs, and anonymous credentials in crowdsensing. In addition, they used
% crowdsourcing to create a machine learning model for parking recognition
% in the presence of foggy filters. Their paper included proof-of-concept
% prototypes such as a Raspberry Pi system and a mobile app, as well as an
% evaluation study of the machine learning model and the effects of crowdsourcing.

\subsubsection{Assistentes de Privacidade}

Existe uma série de assistentes de privacidade no mercado. Os assistentes de
privacidade têm o objetivo de dar ao utilizador flexibilidade na escolha das
opções de privacidade preferidas nas aplicações disponíveis, a maioria é utilizada
em smartphones, muito poucos são feitos para dispositivos na IoT.

% There exists a number of privacy assistants in the market. Privacy assistants
% have the objective of giving the user flexibility in choosing the preferred
% privacy options in available applications, most are used in smartphones,
% very few are made for devices in the IoT.

O CyLab da Carnegie Mellon University, que é o instituto de investigação de
segurança e privacidade da universidade, começou a desenvolver em 2019 uma
infra-estrutura IoT que pretendia estar livre de fugas de privacidade e de
software coberto pela sua iniciativa ``Secure and Private IoT 2019'', este
projeto enquadrar-se-ia no seu principal tema de investigação, a Confiança.
Neste projeto, iniciaram a conceção de um Assistente de Privacidade Personalizado
(PPA) \cite{ColnagoInforming}, que envolveria o uso de entrevistas semi-estruturadas
com 17 participantes para examinar a perceção dos utilizadores de três hipotéticas
implementações PPA, cada uma das quais potencialmente mais autónoma, ao mesmo
tempo que delineavam as vantagens e desvantagens de cada implementação. As
entrevistas foram divididas em três secções: exploratória, ancoragem e PPA;
enquanto o objetivo da fase exploratória era aprender sobre as atitudes dos
participantes e a compreensão da IoT, a fase de ancoragem visava normalizar
a compreensão básica dos participantes sobre como a IoT funciona. A fim de
levar as pessoas a pensar sobre potenciais preocupações de privacidade no final
da secção de ancoragem, os autores perguntaram aos participantes sobre as suas
opiniões sobre privacidade de dados. Na secção PPA, foi proposta a ideia de
uma PPA para a IoT como um potencial projeto futuro. Os autores esclareceram
que o PPA poderia distinguir entre pedidos de dados ativos, tais como um dispositivo
que pedia informações biométricas do rastreador de saúde do utilizador, e recolha
de dados passivos, tais como um dispositivo inteligente com um microfone que
poderia gravar as afirmações das pessoas enquanto estas estivessem por perto.
A Notificação, Recomendação, e Implementação automática de uma PPA IoT foram
as três que os autores e participantes discutiram. Os PPA de Notificação podem
determinar que dispositivos adjacentes estão a solicitar dados e alertar os
utilizadores para a presença e pedidos desses dispositivos, para que os utilizadores
possam aprovar ou rejeitar cada pedido. Com base nos PPA de notificação, os
PPA de recomendação oferecem aos consumidores conselhos sobre como partilhar
os seus dados com base nas suas preferências. As decisões de partilha de dados
dos utilizadores seriam tomadas pelos PPA automáticos. Isto diminuiria a carga
cognitiva dos consumidores, mas também retiraria a sua capacidade de influenciar
o processo. Verificaram que as atitudes dos participantes em relação às várias
implementações eram geralmente favoráveis, embora também manifestassem preocupações,
que variavam consoante o grau de automatização. Dadas as motivações divergentes
dos participantes, alguns desejavam um maior controlo, enquanto outros desejavam
evitar ser ultrapassados por notificações e pela falta de acordo relativamente
à implementação ótima dos PPA.

% The Carnegie Mellon University CyLab, which is the university's security
% and privacy research institute, started developing in 2019 an IoT Infrastructure
% that intended to be free of privacy leaks and software covered by their
% Secure and Private IoT Initiative 2019, this project would fall under
% their main research theme of Trust. In this project they started the design
% of a Personalized Privacy Assistant (PPA) \cite{ColnagoInforming}, this
% would involve the use of semi-structured interviews with 17 participants
% to examine user perceptions of three hypothetical PPA implementations,
% each of which is potentially more autonomous, while outlining the advantages
% and disadvantages of each implementation. The interviews were divided into
% three sections: exploratory, anchoring and the PPA; While the exploratory
% phase's purpose was to learn about participants' attitudes and understanding
% of IoT, the anchoring phase aimed to normalize participants' basic understanding
% of how IoT functions. In order to get people to think about potential privacy
% concerns towards the end of the anchoring section, the authors asked participants
% about their opinions on data privacy. In the PPA section, it was proposed
% the idea of a PPA for IoT as a potential future project. The authors clarified
% that the PPA could distinguish between active data requests such as a gadget
% asking biometric information from the user's health tracker and passive
% data collection such as a smart device with a microphone that could record
% people's utterances while they were nearby. The Notification, Recommendation,
% and Auto implementations of an IoT PPA were the three that the authors and
% attendees discussed. Notification PPAs can determine which adjacent devices
% are requesting data and alert users to those devices' presence and requests
% so that users can approve or reject each request. Building on notification
% PPAs, recommendation PPAs offer consumers advice on how to share their data
% based on their preferences. The user's data sharing decisions would be made
% by auto PPAs. This would lessen the cognitive load on consumers but also
% take away their ability to influence the process. They found that the participants'
% attitudes regarding the various implementations were generally favourable,
% although they also voiced worries, which varied depending on the degree
% of automation. Given the divergent motivations of participants some desired
% increased control, while others wished to avoid being overtaken by notifications
% and the lack of agreement regarding the optimal PPA implementation.

Após a fase de conceção, o instituto implementou um assistente de privacidade
(PA) \cite{FengDesign}, os autores chamaram-lhe IoT Assistant. Devido à
abordagem predominante de ``aviso e escolha'' para a proteção da privacidade
dos dados, decidiram que o PA também cairia nesta abordagem, mas como muitos
sistemas implementam o aviso como forma de consentimento, sem por vezes
oferecerem escolhas ao utilizador final, queriam também que este trabalho
fornecesse um quadro conceptual que visse uma escolha de privacidade centrada
no utilizador, bem como uma taxonomia para os profissionais a utilizar na
conceção de escolhas de privacidade significativas para os seus sistemas.
Os autores definem escolhas de privacidade significativas como ``as capacidades
fornecidas pelos sistemas digitais para os utilizadores controlarem diferentes
práticas de dados sobre os seus dados pessoais'', alargam a noção de escolhas
de privacidade com cinco facetas: eficácia (a oportunidade de estabelecer
preferências de privacidade que correspondam precisa e completamente à recolha
de dados e métodos de utilização que um utilizador está de acordo), eficiência
(a capacidade de especificar estas opções com o mínimo esforço e tempo),
consciência do utilizador (em que as opções de privacidade significativas devem
ser comunicadas de forma proeminente e clara aos utilizadores), abrangência (os
utilizadores devem compreender as suas opções, como afetam a recolha e a
potencial utilização dos seus dados, bem como que conclusões se podem tirar
destes dados e as repercussões potenciais das mesmas) e neutralidade (as decisões
de privacidade significativas não devem ser sujeitas a manipulação ou parcialidade).
O IoT Assistant oferece quatro configurações de privacidade, dando aos
utilizadores finais uma variedade de alternativas para melhor se adequarem
às suas variadas preferências de privacidade e, como resultado, as opções
de privacidade são mais eficazes no ambiente IoT. O IoT Assistant funciona
como uma plataforma centralizada de escolha de privacidade, implementando
várias opções de privacidade, permitindo aos consumidores governar mais
eficazmente a sua privacidade de dados em IoT. Os três modos de descoberta
do sistema IoT que o IoT Assistant suporta são códigos QR, notificações
push, e interfaces de mapa baseadas em localização. Estas ferramentas de
descoberta vão provavelmente tornar os utilizadores mais conscientes dos
dispositivos IoT instalados e das opções de privacidade que têm. Além disso,
o ponto de vista unido da notificação integrada e da opção no IoT Assistant
dá informações sucintas mas completas sobre as práticas de dados IoT para ajudar
os utilizadores a compreender melhor as implicações das suas escolhas de
privacidade. Além disso, os autores trabalham para implementar a notificação
integrada e a opção no IoT Assistant sem preconceitos ou enquadramentos,
tentando oferecer aos consumidores um espaço neutro para executar as suas
escolhas de privacidade. Embora os autores encarem o IoT Assistant como
um passo significativo no sentido de ``opções de privacidade significativas''
na IoT, este assistente ainda tem muitos problemas, tais como o facto de
ainda se encontrar nas suas fases iniciais de desenvolvimento e de não ter
havido muito crescimento dado que foi criado em 2020 e nós estamos em 2023.
Talvez a principal razão pela qual esta aplicação não foi capaz de ser
desenvolvida mais é que a própria aplicação serve para mostrar ao utilizador
os dados que já se encontram na infra-estrutura da IoT que foi criada
anteriormente, e como tal não é capaz de identificar novos dispositivos
IoT sem que os próprios utilizadores finais criem na página web principal \cite{DasPersonalized}
da infra-estrutura uma nova entrada para o dispositivo em questão com o
qual o utilizador quer interagir. Outra razão que paralisa esta aplicação,
bem como outras que procuram proporcionar melhor privacidade nos sistemas de
IoT é que muitos sistemas não oferecem qualquer tipo de escolhas de privacidade
ao utilizador final ou a outros utilizadores que não são os utilizadores
finais pretendidos, mas sobre os quais os dispositivos ainda estão a recolher
dados.

% After the design phase, the institute implemented a privacy assistant (PA) \cite{FengDesign},
% the authors called it IoT Assistant. Because the predominant approach of
% "notice and choice" for data privacy protection, they decided the PA would
% also fall into this approach, but because many systems implement notice
% as a form of consent, without sometimes offering choices to the end user,
% they also wanted this work to provide a conceptual framework that views
% user-centred privacy choice as well as a taxonomy for practitioners to
% use when designing meaningful privacy choices for their systems. The authors
% define meaningful privacy choices as ``the capabilities provided by digital
% systems for users to control different data practices over their personal
% data'', They extend the notion of privacy choices with five facets: effectiveness
% (the opportunity to establish privacy preferences that precisely and completely
% match the data collection and use methods that a user is okay with), efficiency
% (the capacity to specify these options with the least amount of effort and
% time), user awareness (where significant privacy options should be prominently
% and clearly communicated to users), comprehensiveness (users should understand
% their options, how they affect the gathering and potential use of their
% data, as well as what conclusions might be drawn from this data and the
% potential repercussions of these conclusions) and neutrality (meaningful
% privacy decisions should not be subject to manipulation or bias). The IoT
% Assistant offers four privacy settings, giving end users a variety of alternatives
% to better suit their varied privacy preferences and as a result, privacy
% options are more effective in the IoT environment. The IoT Assistant acts
% as a centralized privacy choice platform by implementing various privacy
% options, allowing consumers to more effectively govern their data privacy
% in IoT. The three IoT system discovery modes that the IoT Assistant supports
% are QR codes, push notifications, and location-based map interfaces. These
% discovery tools are probably going to make users more aware of the installed
% IoT devices and the privacy options they have. Additionally, the united
% viewpoint of the integrated notification and option in the the IoT Assistant
% gives succinct yet thorough information regarding IoT data practices to
% help users better understand the implications of their privacy choices.
% Additionally, the authors work to implement the integrated notice and option
% in the IoT Assistant without bias or framing, attempting to offer consumers
% a neutral space to execute their privacy choices. Although the authors view
% the IoT Assistant as a significant step towards ``meaningful privacy options''
% in IoT, this assistant still has many problems, such as the fact that it
% is still in its early stages of development and that there hasn't been much
% growth given that it was created in 2020 and we are in 2023. Maybe the main
% reason this application was not able to be developed further is that the
% application itself serves to show the user the data that is already in the
% IoT infrastructure that was created before, and as such it is not capable
% of identifying new IoT devices without the end users themselves create on
% the infrastructure's main webpage \cite{DasPersonalized} a new entry for
% the device in question that the user wants to interact with. Another reason
% that cripples this application as well as others that seek to provide better
% privacy in IoT systems is that many systems do not offer any type of privacy
% choices to the end user or to other users that are not the intended end
% users but the devices are still collecting data about.

A infra-estrutura da IoT que foi desenvolvida \cite{DasPersonalized} é
construída sobre um design aberto e distribuído que permite a implementação
e gestão de recursos da IoT a ser realizada por qualquer número de atores.
Parte desta infra-estrutura é a Internet of Things Resource Registry, é uma
plataforma web que permite aos proprietários de recursos declarar não só o
local onde um recurso é implantado, mas também práticas de dados como o(s)
motivo(s) para um determinado processo de recolha de dados, o nível de
detalhe nos dados a serem recolhidos, a retenção, os destinatários dos
dados, e muito mais. Além disso, revela quaisquer definições de privacidade
configuráveis pelo utilizador que possam estar ligadas a um determinado
recurso.

% The IoT infrastructure that was developed \cite{DasPersonalized} is built
% on an open, distributed design that allows for the deployment and management
% of IoT resources to be carried out by any number of actors. Part of this
% infrastructure is the Internet of Things Resource Registry, it is a web
% platform that enables resource owners to declare not only the place where
% a resource is deployed but also data practices like the reason(s) for a
% particular data collecting process, the level of detail in the data being
% gathered, retention, the recipients of the data, and more. Additionally,
% it discloses any user-configurable privacy settings that might be connected
% to a particular resource.

\subsection{Principais Levantamentos}

Existem duas formas principais de proporcionar privacidade nos sistemas de
IoT, através da segurança ou da utilização de avisos de privacidade, outras
formas como através de legislação ou com a criação/utilização de uma estrutura
que proporcione privacidade enquadram-se nestas duas categorias. A maior
parte da literatura assume que segurança e privacidade são sinónimos, por
exemplo \cite{opara2022framework, FabianoInternet, SunSecure}, pelo que a
maior parte das soluções propostas são abrangidas pela privacidade através
da segurança. As soluções propostas que utilizam avisos de privacidade,
como \cite{FengDesign}, são implementadas de uma forma que utiliza outros
dispositivos como smartphones que fornecem os próprios avisos, é difícil
fornecer avisos de privacidade nos próprios dispositivos IoT porque muitos
destes dispositivos não têm ecrã ou o ecrã é demasiado pequeno para fornecer
as informações necessárias ao utilizador. Como ainda não existem normas para
a implementação de avisos de privacidade, e as melhores práticas estão dispersas
por toda a literatura, são na sua maioria implementadas de forma aleatória,
pouca orientação é dada aos designers e programadores sobre como fazer um
desenho de aviso de privacidade que seja suficiente e aceitável para o seu
sistema particular e as suas características. Os designers podem desconhecer
as numerosas possibilidades de criar notificações de privacidade aceitáveis e,
como resultado, não as exploram sistematicamente.

% There are two main ways to provide privacy in IoT systems, through security
% or using privacy notices, other ways like through legislation or with the
% creation/usage of a framework that provides privacy fall into these two
% categories. Most of the literature assumes that security and privacy are
% synonyms, for example \cite{opara2022framework, FabianoInternet, SunSecure},
% and so most of the proposed solutions fall under privacy through security.
% The proposed solutions that use privacy notices, like \cite{FengDesign},
% are implemented in a way that use other devices like smartphones that provide
% the notices themselves, it is hard to provide privacy notices on the IoT
% devices themselves because many of these devices do not have a screen or
% the screen is too small to provide the necessary information to the user.
% Because there are still no standards for implementing privacy notices, and
% best practices are scattered throughout the literature, they are mostly
% implemented haphazardly, little guidance is given to designers and developers
% on how to make a privacy notice design that is sufficient and acceptable
% for their particular system and its features. Designers may be unaware of
% the numerous possibilities for creating acceptable privacy notifications
% and, as a result, do not systematically explore them.

Aleisa e Renaud \cite{aleisa2016privacy} também identificam a segurança e a
consciencialização da privacidade como potenciais soluções para as questões de
privacidade na IoT, mas também identificam a minimização dos
dados, o método de boleia e a introspeção. A minimização de dados implica limitar a
recolha de informações pessoais ao que é absolutamente central e reter os
dados apenas pelo tempo necessário para satisfazer o objetivo dos serviços
da tecnologia \cite{ojDirective281}. A boleia \cite{tang2006putting} é um
método de proteção da privacidade dos utilizadores que divulgam a sua
localização, as aplicações consideram a localização como o objeto da sua
atenção. A troca de fidelidade é eliminada, pois não é importante saber
quem se encontra num determinado local. O método de introspeção \cite{kang2015protection}
examina as ações de VM para salvaguardar adequadamente a informação privada
dos utilizadores. O estado do CPU de cada VM, conteúdos de memória, informação
de rede fornecida pelo hipervisor, e qualquer software malicioso que possa
estar presente na VM são todos recolhidos e analisados. A privacidade dos
consumidores é posta em risco se um dispositivo IoT perder integridade devido
a uma agressão hostil.

% Aleisa and Renaud \cite{aleisa2016privacy} also identify security and privacy
% awareness as potential solutions to privacy issues in IoT, but also identify
% data minimization, hitchhiking and introspection. Data minimization entails
% limiting the collecting of personal information to what is absolutely central
% and retaining the data just for as long as is required to satisfy the goal
% of the technology's services \cite{ojDirective281}. Hitchhiking \cite{tang2006putting}
% is a method of protecting the privacy of users who divulge their location,
% applications regard locations as the object of their attention. The fidelity
% trade-off is removed as it is not important to know who is in a certain
% location. The introspection \cite{kang2015protection} method examines VM
% actions to adequately safeguard users' private information. Every VM's
% CPU status, memory contents, network information provided by the hypervisor,
% and any malicious software that may be present on the VM are all collected
% and analysed. The privacy of consumers is jeopardized if an IoT device
% loses integrity due to a hostile assault.

\section{Desafios de Privacidade}

IoT é uma teia complexa de arquiteturas, aplicações e tecnologias. Em termos
de arquiteturas, pode ser decomposta em três camadas: a camada de perceção,
a camada de rede e a camada de aplicação.

% IoT is a composed of a complex web of architectures, applications and technologies.
% In terms of architectures, it can be decomposed in three layers: the perception
% layer, the network layer and the application layer.

A camada de perceção, também conhecida como camada de sensores, interage
com objetos e componentes físicos através de dispositivos inteligentes
(RFID, sensores, atuadores, e assim por diante). Os seus principais objetivos
são a ligação de objetos à rede IoT e a monitorização, recolha e análise de
informação de estado sobre estas coisas utilizando dispositivos inteligentes
implantados. Esta camada pode muitas vezes não ser fiável, por exemplo com
veículos autónomos onde é difícil ler sinais rodoviários ou prever se certos
objetos são inanimados ou não, mas esta falta de fiabilidade também traz
privacidade mesmo que alguns dos dados possam ser inutilizáveis. O ruído
também pode ser acrescentado a esta camada para proporcionar privacidade
extra.

% The perception layer, also known as the sensor layer, interacts with physical
% objects and components via smart devices (RFID, sensors, actuators, and
% so on). Its key objectives are to connect objects to the IoT network and
% to monitor, collect, and analyze status information about these things using
% deployed smart devices. This layer can often be unreliable, for instance
% with autonomous vehicles where they find it hard to read road signs or to
% predict if certain objects are inanimate or not, but this unreliability
% also brings privacy even though some of the data might be unusable. Noise
% can also be added in this layer to provide extra privacy.

Na camada de rede existem muitas redes concorrentes como ZigBee, Z-Wave,
Bluetooth Low Energy, LoRa, Wi-fi, etc., esta camada está fragmentada
especialmente no que diz respeito a redes sem fios e isso torna muito
difícil criar uma arquitetura IoT que possa utilizar várias redes e
ter os vários dispositivos a comunicar entre si, apesar de a interoperabilidade
ser vista como um factor muito importante na IoT. Algumas destas redes
são protocolos padrão abertos enquanto outras são proprietárias e utilizam
diferentes protocolos de comunicação, utilizam diferentes frequências,
diferentes gamas e diferentes taxas de dados. Ao criar uma arquitetura
IoT, os designers pensam frequentemente em como resolver problemas específicos
e utilizar o que é melhor para as necessidades atuais, e a forma como a
IoT é fragmentada não ajuda a proporcionar progresso.

% In the network layer there are many competing networks like ZigBee, Z-Wave,
% Bluetooth Low Energy, LoRa, Wi-fi, etc., this layer is fragmented specially
% in regards to wireless networks and that makes it very difficult to create
% an IoT architecture that can use various networks and have the various
% devices communicate with each other, even though interoperability is seen
% as a very important factor in IoT. Some of these networks are open standard
% protocols while others are proprietary and use different protocols of communication,
% use different frequencies, different ranges and different data rates. When
% creating an IoT architecture the designers often think of how to solve
% specific problems and use what is best for the current needs, and the way
% that IoT is fragmented doesn't help in providing progress.

A camada de aplicação recebe dados da camada de rede e utiliza-os para
executar serviços ou operações essenciais. Esta camada, por exemplo,
pode fornecer o serviço de armazenamento para fazer o backup dos dados
recebidos numa base de dados ou o serviço de análise para analisar os
dados recebidos, a fim de prever o estado futuro dos dispositivos físicos.
Esta camada engloba uma vasta gama de aplicações, cada uma com o seu
próprio conjunto de requisitos. Alguns exemplos são redes inteligentes,
transporte inteligente, e cidades inteligentes.

% The application layer receives data from the network layer and uses it to
% execute essential services or operations. This layer, for example, can provide
% the storage service to backup incoming data into a database or the analysis
% service to analyze received data in order to predict the future state of
% physical devices. This layer encompasses a wide range of applications, each
% with its own set of requirements. A few examples are smart grids, smart
% transportation, and smart cities.

De acordo com Qu et al. \cite{Qu2018Privacy}, permanecem várias barreiras
significativas, incluindo a falta de uma base teórica, a otimização do
trade-off entre privacidade e valor dos dados, e a sobre-complexidade de
isomerismo do sistema. Uma vez que não existem fundamentos matemáticos para
a conceção da estrutura da IoT, os projetos do sistema IoT são planeados
e executados utilizando abordagens empíricas, que têm limitações no desenvolvimento
da IoT. A teoria científica e a análise quantitativa devem permitir uma otimização
trade-off, no entanto, existem múltiplas partes com características e requisitos
diversos, tornando esta otimização altamente desafiante. Uma pletora de normas
e protocolos aumenta a complexidade desnecessária do isomerismo do sistema.
Assegurar aplicações IoT eficazes enquanto desperdiça tão poucos recursos
quanto possível implica menos recursos disponíveis para a proteção da
privacidade, contudo, a proteção da privacidade leve não pode preencher
todos os critérios, e os atacantes podem explorar informação estrutural
para lançar vários ataques simultâneos.

% According to Qu et al. \cite{Qu2018Privacy}, several significant barriers
% remain, including the lack of a theoretical foundation, the trade-off optimization
% between privacy and data value, and system isomerism over-complexity. Because
% there are no mathematical foundations for IoT structure design, IoT system
% designs are planned and executed using empirical approaches, which have
% limitations in IoT development. Scientific theory and quantitative analysis
% must enable trade-off optimization, yet, there are multiple parties with
% diverse characteristics and requirements, making this optimization highly
% challenging. A plethora of standards and protocols add to the unneeded complexity
% of system isomerism. Ensuring effective IoT applications while wasting as
% little resources as feasible implies less resources available for privacy
% protection, however, lightweight privacy protection cannot fulfil all of
% the criteria, and attackers can exploit structural information to launch
% several concurrent attacks.

\section{Metodologia}

O trabalho global será composto por duas fases que serão descritas nos
parágrafos seguintes. A primeira fase, descrita principalmente ao longo
deste trabalho, centra-se na recolha do estado da arte em termos dos tópicos
mais relevantes, dos quais foram selecionados os principais conceitos de
privacidade a serem explorados na fase 1 da fase 2, com a preparação de
um questionário para recolher as perceções dos utilizadores relativamente
à privacidade e aos tópicos recolhidos na revisão sistemática da literatura.
A segunda fase da Fase 2 consiste em desenvolver uma aplicação, parcialmente
baseada na informação gerada pelo inquérito, que possa identificar que tipo
de dispositivos existem, que tipo de dados são recolhidos por esses
dispositivos, apresentar opções de privacidade ao utilizador quando
disponíveis, e o que pode ser feito para evitar que dados indesejáveis
sejam recolhidos.

% The overall work will be comprised of two phases which will be described
% in the following paragraphs. Phase one mainly described throughout this
% paper, focuses on collecting the state of the art in terms of the most relevant
% topics, from which main privacy concepts were selected to be explored in
% the stage 1 of Phase 2 with the preparation of a questionnaire to collect
% user perceptions regarding privacy and topics collected in the systematic
% literature review. The second stage of Phase 2 consists in developing an
% application, partially based on the information generated by the survey,
% that can identify what sort of devices are around, what kind of data is
% gathered by these devices, present privacy options to the user when available,
% and what can be done to prevent undesirable data from being collected.
\par
A Revisão Sistemática da Literatura da Fase 1 reuniu os artigos mais relevantes,
discutindo metodologias e técnicas para a proteção dos dados de privacidade
dos utilizadores, com especial enfoque nos sistemas IoT. Para este SLR,
este artigo considerou focar apenas os artigos dos últimos 12 anos, de 2010
até 2022, uma vez que os artigos antes dessa data se tornaram desactualizados
com a evolução da tecnologia. Neste SLR, foram revistos 54 artigos publicados
nos principais meios informáticos, de segurança, privacidade e engenharia de
software.

% The Phase 1 Systematic Literature Review gathered the most relevant papers
% discussing methodologies and techniques for the protection of users' privacy
% data with special focus on IoT systems. For this SLR, this paper considered
% focusing only on papers from the last 12 years, from 2010 until 2022, since
% papers before then become out of date with the evolution of technology.
% In this SLR, it was reviewed 54 papers published in top computer science,
% security, privacy and software engineering outlets.

Este artigo seguiu a abordagem de Keshav de três passagens \cite{KeshavHow}
ao escolher quais os artigos a ler na íntegra e quais os a ignorar, primeiro
o título seria lido, depois o abstracto, a introdução e a conclusão e, em
seguida, uma breve passagem pelo resto do artigo e depois decidir se valia
a pena ler mais, o ponto focal nesta fase era responder à seguinte pergunta:
o artigo apresenta uma nova metodologia ou um ângulo interessante para abordar
as preocupações dos utilizadores com a privacidade? Só então o documento seria
lido na sua totalidade, ignorando quaisquer tabelas, figuras, imagens ou gráficos.
Se o documento não apresentasse qualquer ideia, abordagem ou técnica interessante,
seria descartado, mas se não o fosse, seria lido cuidadosamente desde o início,
a fim de se compreender plenamente o que apresenta. Tendo recolhido as principais
descobertas, este trabalho visa então conduzir uma divisão ao longo do estudo em
várias fases e em torno das questões específicas de investigação que serão exploradas
em cada fase. Para esse efeito, as questões de investigação enumeradas são:

% This paper followed Keshav's three-pass approach \cite{KeshavHow} when choosing
% which papers to read fully and which ones to ignore, first the title would
% be read, then the abstract, the introduction and conclusion and briefly
% skim the rest of the paper and then decide if it was worth reading any further,
% the focal point in this phase was answering the following question: does
% the paper present a new methodology or interesting angle to tackle users'
% privacy concerns? Only then the document would be read in its entirety while
% ignoring any tables, figures, images or graphs. If the paper failed to present
% any interesting idea, approach, or technique it would be discarded, but
% if not, it would be read carefully from the beginning again in order to
% fully understand what it presents. Having collected the major findings,
% this work then aims to conduct a throughout study split in several stages
% and around the specific research questions which will be explored in each
% phase. For that matter, the research questions listed are:

\vspace{5mm}
\textbf{Fase 1:} \\

% \textbf{Phase 1:} \\

\textbf{RQ1:} Que abordagens estão a ser consideradas para as questões de
privacidade em IoT na literatura atualmente disponível?

% \textbf{RQ1:} What approaches are being considered for privacy issues in
% IoT in the currently available literature?

\textbf{RQ2:} Quais são as percepções dos utilizadores sobre a privacidade
online? \\

% \textbf{RQ2:} What are user perceptions on online privacy? \\

\textbf{Fase 2:} \\

% \textbf{Phase 2:} \\

\textbf{RQ3:} Como capacitar os utilizadores para protegerem os seus direitos
à privacidade?

% \textbf{RQ3:} How to empower users to protect their privacy rights?

\textbf{RQ4:} Que questões prevalecem na IoT que tornam difícil resolver
problemas de privacidade e segurança?

% \textbf{RQ4:} What issues are prevalent in IoT that make it difficult to
% address privacy and security problems?
\vspace{5mm}

A segunda fase será avaliada em duas fases, a primeira consiste em fazer
um estudo sobre as preocupações gerais das pessoas com a privacidade, enquanto
utilizam e interagem com dispositivos IoT. Este estudo será baseado na preparação
de um questionário para avaliar os conhecimentos gerais dos utilizadores
sobre conceitos de privacidade, os seus hábitos e preocupações, a sua compreensão
dos direitos de privacidade, e o que fazem para salvaguardar esses direitos.
O objectivo deste estudo é compreender o paradoxo da privacidade e recolher
dados sobre a sua proposta para abordar questões de privacidade no que diz
respeito aos dispositivos IoT.

% The second phase will be evaluated on two stages, the first one consists
% on doing a study on people's general privacy concerns, while using and interacting
% with IoT devices. This study will abide on preparing a questionnaire to
% assess general user's knowledge on privacy concepts, their habits and concerns,
% their understanding of privacy rights, and what they do to safeguard those
% rights. The goal of this study is to both understand the privacy paradox
% and collect data on their proposal to address privacy issues with regard
% to IoT devices.

\subsection{Estágio 1: Perceções do Utilizador}

Este estudo visa compreender a percepção das pessoas sobre a IoT e as suas
práticas de privacidade em linha. Serve também para desmistificar o paradoxo
da privacidade e também para ajudar a fornecer uma solução para a questão
da privacidade na IoT. O questionário consiste em 92 perguntas divididas em
7 secções para aceder aos conhecimentos dos utilizadores, segue uma espécie
de narrativa, sendo a primeira secção questões gerais de privacidade depois
sobre a predisposição para a partilha de dados, preocupações com a privacidade
depois sobre rotinas digitais diárias, depois sobre identificação de perfis,
e depois sobre o conhecimento geral da Internet das coisas antes de uma secção
sobre dados demográficos não identificáveis. A escala que é utilizada no questionário
baseia-se no trabalho de Philip K. Masur \cite{masur2018situational}. É tomado
grande cuidado quando se trata da recolha de dados deste inquérito, a fim de
não identificar qualquer indivíduo ou grupo de indivíduos, por exemplo, quando
se trata de privacidade diferencial, quaisquer dados que possam identificar
alguém não serão divulgados, mesmo que os dados possam sofrer de alguma imprecisão
devido a isto.

% This study aims to understand people's perception of IoT and their privacy
% practices online. It also serves to demystify the privacy paradox and also
% to help provide a solution to the privacy issue in IoT. The questionnaire
% consists of 92 questions divided into 7 sections to access users' knowledge,
% it follows a kind of narrative, the first section being general privacy
% questions then about the predisposition to data sharing, to concerns with
% privacy then about daily digital routines, then about profile identification,
% and then about IoT general knowledge before a section about non-identifiable
% demographic data. The scale that is used in the questionnaire is based on
% the work of Philip K. Masur \cite{masur2018situational}. Great care is taken
% when it comes to this survey's data collection, in order to not identify
% any individual or group of individuals, for instance, when it comes to differential
% privacy, any data that might identify someone will not be disclosed, even
% though the data might suffer from some inaccuracy because of this.

Este inquérito foi parcialmente baseado num estudo realizado nas Filipinas
pelo governo no contexto da sua lei de privacidade de 2012 \cite{Philippine2022Conduct},
este foi o segundo inquérito realizado sobre a população do país. Foi também
inspirado pela tese de mestrado de Alves \cite{alves2021}, que tratava da
percepção dos cidadãos sobre a privacidade na esteira da GDPR.

% This survey was partially based in a study done in the Philippines by the
% government in the context of their privacy act of 2012 \cite{Philippine2022Conduct},
% this was the second survey done on the country's population. It was also
% inspired by Alves's master's thesis \cite{alves2021}, which was about citizen's
% perception about privacy in the wake of GDPR.

Este inquérito foi feito através da Internet, foi criado em Google Forms,
desta forma é garantido alcançar o maior número de pessoas possível, para
além do Google Formulários propriamente ditos, serão utilizados outros locais
de distribuição em linha e até impressão.

% This survey was done through the internet, it was created in Google Forms,
% this way it is guaranteed to reach the most people possible, besides Google
% Forms itself, it will be used other online venues for distribution and even
% printing.

\subsection{Estágio 2: Estudo no Contexto, uma Aplicação}

Este trabalho propõe uma aplicação que dá aos utilizadores informações sobre
os dispositivos IoT no seu ambiente, como o tipo de informação que estes
dispositivos recolhem e que opções de privacidade estão disponíveis. Esta
aplicação será desenvolvida para telemóveis porque é o dispositivo mais
utilizado que as pessoas levam para onde quer que vão, e porque a aplicação
utilizará georreferenciação para mostrar a localização dos dispositivos IoT.
O principal objectivo desta aplicação é dar aos utilizadores outra opção,
a fim de proteger os seus dados privados. A aplicação mostrará a geolocalização
dos dispositivos IoT, que tipo de dispositivo é, que tipo de dados estão a
ser recolhidos pelo dispositivo. A aplicação não detectará os dispositivos
por si só, isto será feito pelos próprios utilizadores, nas primeiras iterações
desta aplicação foi proposto que a própria aplicação detectaria os dispositivos
e categorizaria que tipo de dispositivo era e que tipo de dados estava a recolher,
mas descobriu-se que esta abordagem era demasiado complexa e, por isso, não
era exequível fazer com os constrangimentos deste documento. A aplicação será
desenvolvida com Flutter, outras opções poderiam ser React Native ou uma aplicação
web progressiva, mas Flutter usa antes do tempo e compilação just in time
com Dart, uma vez que é linguagem de programação, enquanto React Native usa
a linguagem de programação Javascript que nunca foi criada para programação
móvel, pelo que usa uma ponte para converter Javascript em componentes nativos
para Android ou iOS. Flutter tem melhor desempenho e, como tal, é a estrutura
escolhida para esta aplicação.

% This work proposes an application that gives users information about IoT
% devices in their surroundings like the type of information these devices
% collect and what privacy options are available. This application will be
% developed for mobile phones because it is the most used device that people
% take everywhere they go, and because the application will use georeferencing
% to show the location of the IoT devices. The main objective of this application
% is to give users another option in order to protect their private data.
% The application will show the geolocation of the IoT devices, what type
% of device it is, what type of data is being collect by the device. The application
% will not detect the devices by itself, this will be done by the users themselves,
% in the first few iterations of this application it was proposed that the
% application itself would detect the devices and would categorize what type
% of device it was and what type of data it was collecting but it was discovered
% that this approach was too complex and so it was not feasible to do with
% the constraints of this paper. The application will be developed with Flutter,
% other options could be React Native or a progressive web application, but
% Flutter uses ahead of time and just in time compilation with Dart as it
% is programming language while React Native uses the Javascript programming
% language that was never created for mobile programming so it uses a bridge
% to convert Javascript to native components for Android or iOS. Flutter has
% better performance and as such it is the chosen framework for this application.

\section{Estado Atual do Projeto}

Os resultados preliminares do estudo, baseados em 10 respostas, mostram que
todos concordam que a privacidade é importante para eles e algumas pessoas
sabem que não devem partilhar as suas informações pessoais com ninguém em
quem não confiam (como clicar em urls aleatórias ou utilizar websites/software
desprotegidos), mas a maioria delas pensa que a privacidade e a segurança
são o mesmo conceito, a maioria dos inquiridos também não lê avisos de privacidade
mas aceita-os para aceder à informação que querem obter, a maioria dos inquiridos
utiliza os seus dispositivos principalmente para aceder a redes sociais e para
o trabalho, quando se trata de IoT, há uma dissonância entre conhecer o termo
e utilizar dispositivos como relógios inteligentes ou dispositivos activados
por RFID, dos inquiridos que responderam sim à utilização de dispositivos
IoT mais utilizados por causa do trabalho. Note-se também que a maioria dos
inquiridos tem formação em engenharia, pelo que as respostas são enviesadas.
Como resultado, o inquérito permanecerá aberto para reunir um maior número
de respostas e participantes para resultados mais significativos e generalizações.

% The preliminary results of the study, based on 10 responses, show that everyone
% agrees that privacy is important to them and some people know that they
% should not share their personal information with anyone they do not trust
% (like clicking on random urls or using unprotected websites/software), but
% most of them think that privacy and security are the same concept, most
% respondents also do not read privacy notices but accept them to access the
% information they want to get to, most respondents use their devices mostly
% to access social networks and for work, when it comes to IoT, there is a
% dissonance between knowing the term and using devices like smart watches
% or RFID enabled devices, from the respondents that answered yes to using
% IoT devices most use because of work. It is also noted that most respondent
% have a background in engineering, so the responses are skewed. As a result,
% the survey will remain open to gather a larger number of responses and participants
% for more significant results and generalizations.

\begin{table}[ht]
\centering
\begin{adjustbox}{width=0.5\textwidth}
\small
\noindent\begin{tabular}{p{0.17\textwidth}*{20}{|p{0.01\textwidth}}|}
\hline
\multicolumn{0}{|c|}{Plano de trabalho}
    & \multicolumn{4}{c|}{Janeiro}
    & \multicolumn{4}{c|}{Fevereiro}
    & \multicolumn{4}{c|}{Março}
    & \multicolumn{4}{c|}{Abril}
    & \multicolumn{4}{c|}{Maio}
    \\
\hline
\hline
\multicolumn{0}{|l|}{Semana}
    & 1 & 2 & 3 & 4 & 1 & 2 & 3 & 4& 1 & 2 & 3 & 4& 1 & 2 & 3 & 4& 1 & 2 & 3 & 4 \\
\hline
\multicolumn{0}{|l|}{Descoberta e planeamento}
    & \cellcolor[cmyk]{1,1,0,0}&&&& &&&& &&&& &&&&&&& \\
\hline
\multicolumn{0}{|l|}{Inquérito de investigação}
    & \cellcolor[cmyk]{1,1,0,0} & \cellcolor[cmyk]{1,1,0,0} & \cellcolor[cmyk]{1,1,0,0} & \cellcolor[cmyk]{1,1,0,0} & \cellcolor[cmyk]{1,1,0,0} & \cellcolor[cmyk]{1,1,0,0} & \cellcolor[cmyk]{1,1,0,0} & \cellcolor[cmyk]{1,1,0,0} & \cellcolor[cmyk]{1,1,0,0} & \cellcolor[cmyk]{1,1,0,0} & \cellcolor[cmyk]{1,1,0,0} & \cellcolor[cmyk]{1,1,0,0} &&&& &&&& \\
\hline
\multicolumn{0}{|l|}{Estado da arte}
    & \cellcolor[cmyk]{1,1,0,0} & \cellcolor[cmyk]{1,1,0,0} & \cellcolor[cmyk]{1,1,0,0} & \cellcolor[cmyk]{1,1,0,0} & \cellcolor[cmyk]{1,1,0,0} & \cellcolor[cmyk]{1,1,0,0} & \cellcolor[cmyk]{1,1,0,0} & \cellcolor[cmyk]{1,1,0,0} &&&& &&&& &&&& \\
\hline
\multicolumn{0}{|l|}{Requisitos do projecto}
    & \cellcolor[cmyk]{1,1,0,0} & \cellcolor[cmyk]{1,1,0,0} & \cellcolor[cmyk]{1,1,0,0} & \cellcolor[cmyk]{1,1,0,0} &&&& &&&& &&&& &&&& \\
\hline
\multicolumn{0}{|l|}{Wireframes e user stories}
    &&&& & \cellcolor[cmyk]{1,1,0,0} & \cellcolor[cmyk]{1,1,0,0} & \cellcolor[cmyk]{1,1,0,0} & \cellcolor[cmyk]{1,1,0,0} &&&& &&&& &&&& \\
\hline
\multicolumn{0}{|l|}{Prototipagem e refinamento}
    &&&&&& && \cellcolor[cmyk]{1,1,0,0} & \cellcolor[cmyk]{1,1,0,0} & \cellcolor[cmyk]{1,1,0,0} & \cellcolor[cmyk]{1,1,0,0} &&&& &&&& &  \\
\hline
\multicolumn{0}{|l|}{Desenvolvimento}
    &&&& &&&& & \cellcolor[cmyk]{1,1,0,0} & \cellcolor[cmyk]{1,1,0,0} & \cellcolor[cmyk]{1,1,0,0} & \cellcolor[cmyk]{1,1,0,0} & \cellcolor[cmyk]{1,1,0,0} & \cellcolor[cmyk]{1,1,0,0} & \cellcolor[cmyk]{1,1,0,0} & \cellcolor[cmyk]{1,1,0,0} & \cellcolor[cmyk]{1,1,0,0} & \cellcolor[cmyk]{1,1,0,0} & \cellcolor[cmyk]{1,1,0,0} & \cellcolor[cmyk]{1,1,0,0} \\
\hline
\multicolumn{0}{|l|}{Testes e iterações}
    &&&& &&&& &&& & \cellcolor[cmyk]{1,1,0,0} & \cellcolor[cmyk]{1,1,0,0} & \cellcolor[cmyk]{1,1,0,0} & \cellcolor[cmyk]{1,1,0,0} & \cellcolor[cmyk]{1,1,0,0} &&&&  \\
\hline
\multicolumn{0}{|l|}{Lançamento e documentação}
    &&&& &&&& &&&& &&&& & \cellcolor[cmyk]{1,1,0,0} & \cellcolor[cmyk]{1,1,0,0} & \cellcolor[cmyk]{1,1,0,0} & \cellcolor[cmyk]{1,1,0,0} \\
\hline
\end{tabular}
\end{adjustbox}
\vspace{1em}
\caption{Work plan timeline}
\label{workchart}
\end{table}

Como se pode ver na Tabela \ref{workchart}, os primeiros meses envolverão
a concepção da aplicação e o inquérito do estudo seguido do desenvolvimento
da aplicação e da síntese do estudo, e finalmente o teste e refinamento da
aplicação. Devido ao carácter exploratório deste trabalho, a aplicação poderá
sofrer alterações no desenho, especialmente na fase de teste, e também em
função dos resultados do estudo.

% As can be seen in Table \ref{workchart}, the first months will involve the
% design of the application and the enquiry of the study followed by the development
% of the application and the synthesis of the study, and finally testing and
% refinement of the application. Because of the exploratory nature of this
% work the application might suffer alterations to the design, specially in
% the testing stage, and also depending on the results of the study.

\section{Trabalho Futuro}

Embora existam soluções de hardware que podem detectar alguns dispositivos
em redes particulares, como ZigBee ou Bluetooth LE, nomeadamente sniffers
IoT e existam algumas aplicações de georeferenciação que tentam localizar
certos dispositivos IoT, ainda é necessário algum tipo de dispositivo ou
estrutura que seja agnóstico de rede e possa detectar onde os dispositivos
estão localizados e que tipo de dados os dispositivos IoT que estão à sua
volta estão a recolher. Este dispositivo deve também ser capaz de informar os
utilizadores sobre os avisos de privacidade dos dispositivos e o que podem
os utilizadores fazer para salvaguardar os seus dados pessoais. Os sniffers
IoT disponíveis são utilizados principalmente na detecção de problemas na comunicação
de dispositivos na rede ou para resolver problemas de interoperabilidade entre
diferentes redes IoT. Existem muitos obstáculos que impedem a criação de um
tal dispositivo e o facto de ainda não existir nada parecido com o que possa
estar relacionado com o mesmo, ou não existe interesse suficiente por parte
dos utilizadores ou investigadores para se concentrarem num tal esforço ou a
complexidade de tal tarefa é maior do que as recompensas.

% Although there are existing hardware solutions that can detect some devices
% on particular networks, like ZigBee or Bluetooth LE, namely IoT sniffers
% and there exist some georeferencing applications that try to pinpoint certain
% IoT devices, there is still a need for some kind of device or framework
% that is network agnostic and can detect where the devices are located and
% what kind of data the IoT devices that are around it are collecting. This
% gadget should also be capable of informing users about the privacy notices
% of the devices and what can the users do to safeguard their personal data.
% The IoT sniffers that are available are primarily used in the detection
% of problems in the communication of devices in the network or to solve problems
% of interoperability between different IoT networks. There are many obstacles
% that impede the creation of such a device and the fact that it still does
% not exist anything like it may be related ton either there iss not enough
% interest from users or researchers to focus on such an endeavour or the
% complexity of such a task is greater than the rewards.

\section{Conclusão}

Este projecto visa fazer uma análise exploratória da privacidade em sistemas
de IoT. Propõe um inquérito para melhor compreender os conhecimentos dos utilizadores
sobre este assunto e uma aplicação que visa criar mais consciência dos utilizadores
e informar melhor sobre o seu ambiente, bem como sobre os dispositivos da IoT
que o habitam e como podem responder em conformidade.

% This project aims to do an exploratory analysis of privacy in IoT systems.
% It proposes a survey to better understand user's knowledge on this subject
% and an application that aims to create more users awareness and better inform
% about their environment, as well as the IoT devices that inhabit it and
% how they can respond accordingly.

Espera-se que o trabalho realizado neste projecto seja útil para apoiar ainda
mais os investigadores e que a aplicação que será desenvolvida seja capaz de proporcionar
uma maior visibilidade, permitindo assim aos utilizadores adquirir conhecimentos
sobre os dados que estão a ser recolhidos e como podem ajustar o seu comportamento
ou responder mais eficazmente para proteger os seus direitos de privacidade.

% Hopefully the work conducted on this project will be useful to further support
% researchers and the application that will be developed will be able to
% provide greater visibility, thus allowing users to acquire knowledge about
% the data being collected and how they can adjust their behaviour or respond
% more effectively to protect their privacy rights.

\bibliographystyle{IEEEtran}
\bibliography{assets/references}

\end{document}
